---
title: "In this field, insert the project title"
author: "In this field, insert the author(s) of the analysis"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends
---
================================================================================
# Study description:


~ Include here a brief description of the study ~


================================================================================

# OVERVIEW - TIMESERIES

# 1. Load Libraries
```{r libraries_overview, eval=TRUE, echo=FALSE, warning=FALSE, include=FALSE}

library(gtsummary)
library(writexl)
library(dplyr)
library(phyloseq)
library(nlme)
library(vegan)
library(compositions)
library(ggplot2)
library(lme4)
library(reshape2) 
library(vegan)
library(ade4)
library(plotly)
library(pracma) 
library(fpc) 
library(tidyverse)
library(cluster)
library(RColorBrewer)
library(ape)
library(gplots)
library(RColorBrewer)
library(pheatmap)
library(Maaslin2)
library(ggpicrust2)
library(microbiome)
library(file2meco)
library(microbiomeutilities)
library(gridExtra)
library(GGally)
library(broom)
library(broom.mixed)
library(kableExtra)
library(emmeans)
library(cardx)
library(tibble)
library(knitr)
library(forcats)
library(here)

```

# 2. Data preparation

## 2.1. Data preparation and summary data table  
```{r data_prep, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

#  Path
path     <- here::here() 
overview <- file.path(path, "overview")
dir.create(overview, showWarnings = FALSE, recursive = TRUE)

# Load data
ps0 <- readRDS("ps.dna.rds")

# Adjust rownames from sample data to adjust table
samdf <- sample_data(ps0)
otu_table <- otu_table(ps0)
rownames(samdf) <- samdf$SampleID
rownames(otu_table) <- samdf$SampleID 

# organize the ps0 object with update names
ps0 <- phyloseq(tax_table(ps0),
                sample_data(samdf),
                otu_table(otu_table, taxa_are_rows = FALSE))

# Set the reference group 
sample_data(ps0)$Group <- factor(
  sample_data(ps0)$Group,
  levels = c("Group1", "Group2")
)

#  Additional adjustments in ps0 object (df enhancement)
sample_data(ps0)$Animal <- sprintf("A%03d",
                                   as.numeric(gsub("A", "",
                                                   sample_data(ps0)$Animal)))
sample_data(ps0)$Timepoint <- factor(
  sample_data(ps0)$Timepoint,
  levels = c(1, 2, 3),
  labels = c("TP1", "TP2", "TP3")
)

# Extracts and saves adjusted metadata
samdf <- microbiome::meta(ps0)
saveRDS(samdf, file.path(overview, "adjusted_metadata.rds"))
saveRDS(ps0,  "ps0.rds")

# Create table 
summary_table <- samdf %>%
  select(Group, Timepoint) %>%
  tbl_summary(
    by        = Timepoint,
    missing   = "no",
    statistic = all_categorical() ~ "{n}/{N} ({p}%)",
    label     = list(Group ~ "Group")
  ) %>%
  add_p(test = all_categorical() ~ "fisher.test") %>%
  modify_header(
    label          ~ "**Variable**",
    all_stat_cols() ~ "**{level}**"              # {level} became D20, D25, D103
  ) %>%
  modify_spanning_header(
    all_stat_cols() ~ "**Timepoint (N={N})**"    # display N= in each column 
  ) %>%
  bold_labels()

# Export to excel
write_xlsx(
  as_tibble(summary_table$table_body),
  file.path(overview, "summary_statistics.xlsx")
)

# Format table for printing 
print(
  summary_table %>%
    as_kable_extra(
      booktabs    = TRUE,
      linesep     = "",
      caption     = "Descriptive Table by Timepoint"
    ) %>%
    kable_styling(
      latex_options = c("hold_position", "scale_down")
    )
)

# clear environment
#rm(list = ls(all = TRUE))

```
We were able to identify loss to follow-up with substitution, resulting in one fewer sample at Day 20.
\FloatBarrier

## 2.2. Plot Data Preparation
```{r data_prep_plot, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4,  fig.cap='Prevalence vs. Total Abundance (ASVs) with filtering cutoff'}

# Paths
path         <- here::here() 
plots_dir <- file.path(path, 'prev_abund_plots')
dir.create(plots_dir)

# load data
ps.dna <- readRDS("ps0.rds")
ps0.dna <- subset_taxa(ps.dna,
                       !is.na(Phylum) &
                       !Phylum %in% c("", "uncharacterized", "NA"))
prevalenceThreshold <- 0.1 * nsamples(ps0.dna)

# Prevalence x Abundance
prev_counts <- apply(otu_table(ps0.dna) > 0,
                     ifelse(taxa_are_rows(ps0.dna), 1, 2),
                     sum)
abund_counts <- taxa_sums(ps0.dna)
prevdf_asv <- data.frame(Prevalence = prev_counts,
                         Abundance  = abund_counts)

# Plot
prev_plot <- ggplot(prevdf_asv, aes(x = Prevalence, y = Abundance)) +
  geom_point(alpha = 0.4) +
  geom_vline(xintercept = prevalenceThreshold,
             linetype = "dashed", color = "red") +
  labs(
    title    = "Prevalence vs. Total Abundance (ASVs)",
    subtitle = paste0("Filtering cutoff = ", prevalenceThreshold, " samples"),
    x        = "Number of samples in which ASV appears",
    y        = "Total read count"
  ) +
  theme_bw(base_size = 12)

# Save
ggsave(filename = "prevalence_vs_abundance.png" , plot = prev_plot, path = plots_dir, width = 6, height = 4, units = "in", dpi = 300)
ggsave(filename = "prevalence_vs_abundance.pdf" , plot = prev_plot, path = plots_dir, width = 6, height = 4, units = "in", dpi = 300)
saveRDS(ps0.dna, "ps0.dna.rds")

```


## 2.3. Data preparation overview table
```{r data_prep_table, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

# Load data
ps0.dna      <- readRDS("ps0.rds")
ps0.filtered <- subset_taxa(ps0.dna,
                            !is.na(Phylum) &
                            !Phylum %in% c("", "uncharacterized", "NA"))

# Filtering based on prevalence and abundance
prevalenceThreshold <- 0.1 * nsamples(ps0.filtered)                # 10% das amostras
totalReads          <- sum(taxa_sums(ps0.filtered))
abundanceThreshold  <- 0.0001 * totalReads                           # 0.01% do total de reads

# Filtering ASV
keepASV   <- taxa_sums(ps0.filtered) >= abundanceThreshold &
             apply(otu_table(ps0.filtered) > 0,
                   ifelse(taxa_are_rows(ps0.filtered), 1, 2),
                   sum) >= prevalenceThreshold
ps1.dna   <- prune_taxa(keepASV, ps0.filtered)

# Filtering Genus
ps0.genus      <- tax_glom(ps0.filtered, "Genus",  NArm = FALSE)
keepGenus      <- taxa_sums(ps0.genus) >= abundanceThreshold &
                  apply(otu_table(ps0.genus) > 0,
                        ifelse(taxa_are_rows(ps0.genus), 1, 2),
                        sum) >= prevalenceThreshold
ps1.dna.genus  <- prune_taxa(keepGenus, ps0.genus)

# Filtering Family
ps0.family     <- tax_glom(ps0.filtered, "Family", NArm = FALSE)
keepFamily     <- taxa_sums(ps0.family) >= abundanceThreshold &
                  apply(otu_table(ps0.family) > 0,
                        ifelse(taxa_are_rows(ps0.family), 1, 2),
                        sum) >= prevalenceThreshold
ps1.dna.family <- prune_taxa(keepFamily, ps0.family)

# Filtering Phylum
ps0.phyla      <- tax_glom(ps0.filtered, "Phylum", NArm = FALSE)
keepPhylum     <- taxa_sums(ps0.phyla) >= abundanceThreshold &
                  apply(otu_table(ps0.phyla) > 0,
                        ifelse(taxa_are_rows(ps0.phyla), 1, 2),
                        sum) >= prevalenceThreshold
ps1.dna.phy    <- prune_taxa(keepPhylum, ps0.phyla)

# Save
saveRDS(ps1.dna,        "ps1.dna.rds")
saveRDS(ps1.dna.genus,  "ps1.dna.genus.rds")
saveRDS(ps1.dna.family, "ps1.dna.family.rds")
saveRDS(ps1.dna.phy,    "ps1.dna.phy.rds")

# Create table
counts <- tibble(
  Level      = c("ASV", "Genus", "Family", "Phylum"),
  Before     = c(ntaxa(ps0.filtered), ntaxa(ps0.genus),
                 ntaxa(ps0.family),    ntaxa(ps0.phyla)),
  After      = c(ntaxa(ps1.dna),      ntaxa(ps1.dna.genus),
                 ntaxa(ps1.dna.family),ntaxa(ps1.dna.phy)),
  `Prev Thres`  = round(prevalenceThreshold, 1),
  `Abund Thres` = round(abundanceThreshold, 0)
)

# Format table for printing
counts %>%
  kable(
    caption = "Taxa before/after prevalence \\& abundance filtering",
    digits  = 0,
    booktabs= TRUE
  ) %>%
  kable_classic(full_width = FALSE, position = "center") %>%
  column_spec(2:3, width = "1.5cm") %>%
  column_spec(4:5, width = "1.2cm")

# 9) Limpa ambiente
#rm(list = ls(all = TRUE))

```
\FloatBarrier

We filtered out taxa lacking phylum-level annotations, applied a 10% prevalence filter, 0.01% abundance filter, and aggregated taxa at the phylum, family, and genus levels.

# 3. Microbiome overview

## 3.1. Composition of microbiome
```{r dataSummary, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# Paths
path         <- here::here()
overview_dir <- file.path(path, 'overview')
dir.create(overview_dir, showWarnings = FALSE, recursive = TRUE)

# Load data
ps_ASV    <- readRDS(file.path(path, 'ps1.dna.rds'))

# Agglomerate by taxonomic level
ps_phylum <- tax_glom(ps_ASV, 'Phylum', NArm = TRUE)
ps_family <- tax_glom(ps_ASV, 'Family', NArm = TRUE)
ps_genus  <- tax_glom(ps_ASV, 'Genus', NArm = TRUE)

# Metadata // Related info
# Adjust accordingly group names and factors (categorical variables)
meta <- data.frame(sample_data(ps_ASV)) %>%
  mutate(
    Sample      = sample_names(ps_ASV),
    Depth       = sample_sums(ps_ASV),
    Group = factor(Group, levels = c('Group1','Group2')),
    Timepoint = factor(paste0("TP", Timepoint), levels = c("TP1","TP2","TP3"))
  )

# 4) two‑way ANOVA  e tabela tidy
depth_aov <- aov(Depth ~ Group * Timepoint, data = meta)
anova_tbl <- broom::tidy(depth_aov)
colnames(anova_tbl)[1:6] <- c('Term','Df','Sum Sq','Mean Sq','F value','Pr(>F)')

# Format table for printing (kable pack)
anova_tbl %>%
  kable(caption = 'Two‑way ANOVA on Sequencing Depth', digits = 3, booktabs = TRUE) %>%
  kable_styling(latex_options = c('hold_position','scale_down')) %>%
  print()

# 10) Limpa ambiente
#rm(list = ls(all = TRUE))

```
\FloatBarrier

```{r depthSeq, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# Sequencing depth boxplot by group and time point.
p0 <- ggplot(meta, aes(x = Group, y = Depth, fill = Group)) +
  geom_boxplot(width = 0.6) +
  scale_fill_manual(values = c('Group1' = 'steelblue',
                               'Group2' = 'orange'
                               )) +
  facet_grid(. ~ Timepoint) +                         # <- aqui
  labs(title = 'Library size by Group and Timepoint',
       x = 'Group', y = 'Sequencing depth (total number of reads)'
       ) +
  theme_bw() +
  theme(legend.position = 'none')

ggsave(file.path(overview_dir, 'fig0_depth_boxplot.pdf'),
       p0, width = 180, height = 120, units = 'mm', dpi = 300)

print(p0)

# Save
ggsave(
  file.path(overview_dir, 'fig0_depth_boxplot.pdf'),
  p0, width = 180, height = 120, units = 'mm', dpi = 300
)

```

```{r librarySize, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# library size barchart by sample
p1 <- ggplot(meta, aes(x = Sample, y = Depth, fill = Group)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('Group1' = 'steelblue', 
                               'Group2' = 'orange'
                               )) +
  labs(title = 'Library size by Sample', 
       x = 'Sample ID', 
       y = 'Sequencing depth (total number of reads)', 
       fill = 'Group'
       ) +
  theme_bw(base_size = 8) + 
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1), 
        legend.position = 'top')

ggsave(file.path(overview_dir, 'fig1_library_size.pdf'), p1, width = 180, height = 120, units = 'mm', dpi = 300)

print(p1)

```

```{r funcRelat_abundance, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# Function for relative abundance 
taxa_plot <- function(ps_obj, level_col, top_n, fig_num) {
  ps_ra <- transform_sample_counts(ps_obj, function(x) x / sum(x))
  df    <- psmelt(ps_ra)
  # se usar level_col='ASV', renomeia coluna OTU para ASV
  if (level_col == 'ASV') df <- rename(df, ASV = OTU)
  # identifica coluna real para agrupamento
  real_col <- if (level_col == 'ASV') 'ASV' else level_col
  # calcula top taxa
  top_taxa <- df %>%
    group_by(.data[[real_col]]) %>%
    summarize(mean_abund = mean(Abundance), .groups = 'drop') %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = top_n) %>%
    pull(real_col)
  # agrega para plot
  df_plot <- df %>%
    mutate(Taxa = as.character(.data[[real_col]]), Taxa = ifelse(Taxa %in% top_taxa, Taxa, 'Other')) %>%
    group_by(Sample, Group, Timepoint, Taxa) %>%
    summarize(Abundance = sum(Abundance), .groups = 'drop') %>%
    mutate(Taxa = fct_reorder(Taxa, Abundance, .fun = mean))
  
  p <- ggplot(df_plot, aes(x = Sample, y = Abundance, fill = Taxa)) +
    geom_bar(stat = 'identity') +
    facet_wrap(vars(Timepoint, Group), ncol = 2, scales = 'free_x') +
    labs(title = paste0('Figure ', fig_num, ': Relative Abundance at ', level_col),
         x = 'Sample ID', 
         y = 'Relative Abundance', 
         fill = level_col) +
    theme_bw(base_size = 8) + 
    theme(axis.text.x = element_text(angle = 90, 
                                     vjust = 0.5, 
                                     hjust = 1, size = 6),
                        strip.background = element_rect(
                          fill = 'white'),
                        strip.text = element_text(face = 'bold'),
                        legend.position = 'right')
  # save
  fname <- paste0('fig', fig_num, '_', tolower(level_col), '.pdf')
  ggsave(file.path(overview_dir, fname), p, width = 180, height = 120, units = 'mm', dpi = 300)
  
  print(p)
}

```

```{r f4, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at phylum level"}

taxa_plot(ps_phylum, 'Phylum', 10, '2')

```

```{r f5, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at family level"}

taxa_plot(ps_family, 'Family', 10, '3')

```

```{r f6, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at genus level"}

taxa_plot(ps_genus,  'Genus',  20, '4')

```

```{r f7, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at ASV level"}

taxa_plot(ps_ASV,    'ASV',    20, '5')

```

### 3.2. General Dominant taxa
Here we transform the data in relative abundances for proper evaluation.
```{r microbiome_overview, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here::here() 
overview = paste(path,"/overview/",sep = "")

# Load data
ps0.dna <- readRDS("ps1.dna.rds")

# Agglomerating
ps.dna.phy <- tax_glom(ps0.dna, "Phylum", NArm = TRUE)
ps.dna.family <- tax_glom(ps0.dna, "Family", NArm = TRUE)
ps.dna.genus <- tax_glom(ps0.dna, "Genus", NArm = TRUE)

# ASV level
ps.ra <- transform_sample_counts(ps0.dna, function(x) x/sum(x))

# Phylum level
taxa_names(ps.dna.phy) <- tax_table(ps.dna.phy)[,2]
ps.phy.ra <- transform_sample_counts(ps.dna.phy, function(x) x/sum(x))

# Family level
taxa_names(ps.dna.family) <- tax_table(ps.dna.family)[,5]
ps.family.ra <- transform_sample_counts(ps.dna.family, function(x) x/sum(x))

# Genus level
# taxa_names(ps.dna.genus) <- tax_table(ps.dna.genus)[,6]
ps.genus.ra <- transform_sample_counts(ps.dna.genus, function(x) x/sum(x))
genus.melt <- psmelt(ps.genus.ra)

```

The distribution of the reads are here summarized on phylum, family and genus level. 
```{r ov_microbiome_tables, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

df2 <- data.frame(tax_table(ps.dna.phy), taxprc = 100*taxa_sums(ps.phy.ra)/length(sample_names(ps.phy.ra)))
df3 <- data.frame(tax_table(ps.dna.family), taxprc = 100*taxa_sums(ps.family.ra)/length(sample_names(ps.family.ra)))
df4 <- data.frame(tax_table(ps.dna.genus),taxprc = 100*taxa_sums(ps.genus.ra)/length(sample_names(ps.genus.ra)))
df5 <- data.frame(tax_table(ps0.dna),taxprc = 100*taxa_sums(ps.ra)/length(sample_names(ps.ra)))

df.count <- data.frame(Included = c("All", "> 0.01%", "> 0.1%","> 1%"),
                          Phylum = c(nrow(df2),sum(df2$taxprc > 0.01),sum(df2$taxprc > 0.1),sum(df2$taxprc > 1)),
                          Family = c(nrow(df3),sum(df3$taxprc > 0.01),sum(df3$taxprc > 0.1),sum(df3$taxprc > 1)),
                          Genus = c(nrow(df4),sum(df4$taxprc > 0.01),sum(df4$taxprc > 0.1),sum(df4$taxprc > 1)),
                          ASV = c(nrow(df5),sum(df5$taxprc > 0.01),sum(df5$taxprc > 0.1),sum(df5$taxprc > 1)))

# Output file - Export to excel 

# abundance df for different taxonomic levels
FileName <- paste(overview,"/Abundance of phyla, family, genera and ASV in microbiome samples.xlsx", sep = "")
write_xlsx(df.count,FileName)

# abundance genus level df
FileName <- paste(overview,"/Average abundance according to genus.xlsx", sep = "")
write_xlsx(df4,FileName)

# abundance family level df
FileName <- paste(overview,"/Average abundance according to family.xlsx", sep = "")
write_xlsx(df3,FileName)

# abundance phylum level df
FileName <- paste(overview,"/Average abundance according to phylum.xlsx", sep = "")
write_xlsx(df2,FileName)

# Count of ASV and taxa in samples
kable(df.count, row.names = F,digits = 1, caption = 'Abundance of phyla, family, genera and ASV in microbiome samples')%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Top 6 dominating phyla (prc abundance)
kable(head(df2[order(df2$taxprc,decreasing = T),c("Kingdom","Phylum","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to phylum',col.names = c("Kingdom","Phylum","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Top 6 dominating family (prc abundance)
kable(head(df3[order(df3$taxprc,decreasing = T),c("Kingdom","Phylum","Family", "taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to family',col.names = c("Kingdom","Phylum", "Family", "Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Top 6 dominating (genera prc abundance)
kable(head(df4[order(df4$taxprc,decreasing = T),c("Kingdom","Phylum","Class","Order","Family","Genus","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to genus',col.names = c("Kingdom","Phylum","Class","Order","Family","Genus","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# clean environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

### 3.3. Distribution of taxa in the samples of Group 1
```{r group1_dist1, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here:here()
over_group_timepoint = paste(path,"/overview_by_group_and_timepoint/",sep = "")
dir.create(over_group_timepoint)

# Load data
ps <- readRDS("ps1.dna.rds")
ps.group1 <- subset_samples(ps, Group=='Group1') # Adjust group and variable name accordingly

# Agglomerating 
ps.group1.phy <- tax_glom(ps.group1, "Phylum", NArm = TRUE)
ps.group1.family <- tax_glom(ps.group1, "Family", NArm = TRUE)
ps.group1.genus <- tax_glom(ps.group1, "Genus", NArm = TRUE)

# ASV level
ps.group1.ra <- transform_sample_counts(ps.group1, function(x) x/sum(x))

# Phylum level
taxa_names(ps.group1.phy) <- tax_table(ps.group1.phy)[,2]
ps.group1.phy.ra <- transform_sample_counts(ps.group1.phy, function(x) x/sum(x))

# Family level
taxa_names(ps.group1.family) <- tax_table(ps.group1.family)[,5]
ps.group1.family.ra <- transform_sample_counts(ps.group1.family, function(x) x/sum(x))

# Genus level
#taxa_names(ps.dna.genus) <- tax_table(ps.dna.genus)[,6]
ps.group1.genus.ra <- transform_sample_counts(ps.group1.genus, function(x) x/sum(x))
ps.group1.genus.melt <- psmelt(ps.group1.genus.ra)

# Calculating relative abundance
df2 <- data.frame(tax_table(ps.group1.phy), 
                  taxprc = 100*taxa_sums(ps.group1.phy.ra)/
                    length(sample_names(ps.group1.phy.ra)))
df3 <- data.frame(tax_table(ps.group1.family), 
                  taxprc = 100*taxa_sums(ps.group1.family.ra)/
                    length(sample_names(ps.group1.family.ra)))
df4 <- data.frame(tax_table(ps.group1.genus),
                  taxprc = 100*taxa_sums(ps.group1.genus.ra)/
                    length(sample_names(ps.group1.genus.ra)))
df5 <- data.frame(tax_table(ps.group1),
                  taxprc = 100*taxa_sums(ps.group1.ra)/
                    length(sample_names(ps.group1.ra)))

df.count <- data.frame(Included = c("All", "> 0.01%", "> 0.1%","> 1%"),
                          Phylum = c(nrow(df2),sum(df2$taxprc > 0.01),sum(df2$taxprc > 0.1),sum(df2$taxprc > 1)),
                          Family = c(nrow(df3),sum(df3$taxprc > 0.01),sum(df3$taxprc > 0.1),sum(df3$taxprc > 1)),
                          Genus = c(nrow(df4),sum(df4$taxprc > 0.01),sum(df4$taxprc > 0.1),sum(df4$taxprc > 1)),
                          ASV = c(nrow(df5),sum(df5$taxprc > 0.01),sum(df5$taxprc > 0.1),sum(df5$taxprc > 1)))

# Count of ASV and taxa in group 1 samples
kable(df.count, row.names = F,digits = 1, caption = 'Abundance of phyla, family, genera and ASV in Group 1 samples')%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Abundance of phyla, family, genera and ASV in Group 1 samples.xlsx", sep = "")
write_xlsx(df.count,FileName)

# Top 6 dominating phyla (prc abundance)
kable(head(df2[order(df2$taxprc,decreasing = T),c("Kingdom","Phylum","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to phylum in Group 1 samples',col.names = c("Kingdom","Phylum","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Average abundance according to phylum in Group 1 samples.xlsx", sep = "")
write_xlsx(df2,FileName)

# Top 6 dominating family (prc abundance)
kable(head(df3[order(df3$taxprc,decreasing = T),c("Kingdom","Phylum","Family", "taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to family in Group 1 samples',col.names = c("Kingdom","Phylum", "Family", "Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Average abundance according to family in Group 1 samples.xlsx", sep = "")
write_xlsx(df3,FileName)

# Top 6 dominating (genera prc abundance)
kable(head(df4[order(df4$taxprc,decreasing = T),c("Kingdom","Phylum","Class","Order","Family","Genus","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to genus in Group 1 samples',col.names = c("Kingdom","Phylum","Class","Order","Family","Genus","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Average abundance according to genus in Group 1 samples.xlsx", sep = "")
write_xlsx(df4,FileName)

# clean environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

### 3.4. Distribution of taxa in the samples of Group 2
```{r group_2_dist1, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here::here()
over_group_timepoint = paste(path,"/overview_by_group_and_timepoint/",sep = "")

# Load data
ps <- readRDS("ps1.dna.rds")
ps.beta <- subset_samples(ps, Group=='Group2')

#Aglomerações
ps.group2.phy <- tax_glom(ps.group2, "Phylum", NArm = TRUE)
ps.group2.family <- tax_glom(ps.group2, "Family", NArm = TRUE)
ps.group2.genus <- tax_glom(ps.group2, "Genus", NArm = TRUE)

# ASV level
ps.group2.ra <- transform_sample_counts(ps.group2, function(x) x/sum(x))

# Phylum level
taxa_names(ps.group2.phy) <- tax_table(ps.group2.phy)[,2]
ps.group2.phy.ra <- transform_sample_counts(ps.group2.phy, function(x) x/sum(x))

# Family level
taxa_names(ps.group2.family) <- tax_table(ps.group2.family)[,5]
ps.group2.family.ra <- transform_sample_counts(ps.group2.family, function(x) x/sum(x))

# Genus level
# taxa_names(ps.dna.genus) <- tax_table(ps.dna.genus)[,6]
ps.group2.genus.ra <- transform_sample_counts(ps.group2.genus, function(x) x/sum(x))
ps.group2.genus.melt <- psmelt(ps.group2.genus.ra)

df2 <- data.frame(tax_table(ps.group2.phy),
                  taxprc = 100*taxa_sums(ps.group2.phy.ra)/
                    length(sample_names(ps.group2.phy.ra)))
df3 <- data.frame(tax_table(ps.group2.family), 
                  taxprc = 100*taxa_sums(ps.group2.family.ra)/
                    length(sample_names(ps.group2.family.ra)))
df4 <- data.frame(tax_table(ps.group2.genus),
                  taxprc = 100*taxa_sums(ps.group2.genus.ra)/
                    length(sample_names(ps.group2.genus.ra)))
df5 <- data.frame(tax_table(ps.group2),
                  taxprc = 100*taxa_sums(ps.group2.ra)/
                    length(sample_names(ps.group2.ra)))

df.count <- data.frame(Included = c("All", "> 0.01%", "> 0.1%","> 1%"),
                          Phylum = c(nrow(df2),sum(df2$taxprc > 0.01),sum(df2$taxprc > 0.1),sum(df2$taxprc > 1)),
                          Family = c(nrow(df3),sum(df3$taxprc > 0.01),sum(df3$taxprc > 0.1),sum(df3$taxprc > 1)),
                          Genus = c(nrow(df4),sum(df4$taxprc > 0.01),sum(df4$taxprc > 0.1),sum(df4$taxprc > 1)),
                          ASV = c(nrow(df5),sum(df5$taxprc > 0.01),sum(df5$taxprc > 0.1),sum(df5$taxprc > 1)))

# Count of ASV and taxa in Group 2 samples
kable(df.count, row.names = F,digits = 1, caption = 'Abundance of phyla, family, genera and ASV in Group 2 samples')%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Abundance of phyla, family, genera and ASV in Group 2 samples.xlsx", sep = "")
write_xlsx(df.count,FileName)

# Top 6 dominating phyla (prc abundance)
kable(head(df2[order(df2$taxprc,decreasing = T),c("Kingdom","Phylum","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to phylum in Group 2 samples',col.names = c("Kingdom","Phylum","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Average abundance according to phylum in Group 2 samples.xlsx", sep = "")
write_xlsx(df2,FileName)

# Top 6 dominating family (prc abundance)
kable(head(df3[order(df3$taxprc,decreasing = T),c("Kingdom","Phylum","Family", "taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to family in Group 2 samples',col.names = c("Kingdom","Phylum", "Family", "Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Average abundance according to family in Group 2 samples.xlsx", sep = "")
write_xlsx(df3,FileName)

# Top 6 dominating (genera prc abundance)
kable(head(df4[order(df4$taxprc,decreasing = T),c("Kingdom","Phylum","Class","Order","Family","Genus","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to genus in Group 2 samples',col.names = c("Kingdom","Phylum","Class","Order","Family","Genus","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group_timepoint,"/Average abundance according to genus in Group 2 samples.xlsx", sep = "")
write_xlsx(df4,FileName)

# clean environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

# 4. Groups composition 
## 4.1. Top 10 Phyla - Group1 x Group2 composition by timepoint 
```{r phylum_abundance_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Top 10 phylum abundance by group'}

# Paths
path <- here::here()
over_Group_Timepoint = paste(path,"/overview_by_group_and_timepoint/",sep = "")

# Load aglomerate data by phylum 
ps <- readRDS("ps1.dna.phy.rds")

# Transform absolute abundance and calculating relative abundance
ps_ra <- transform_sample_counts(ps, function(x) x / sum(x))
ps_ra_melt <- psmelt(ps_ra)

# Color palette
palette <- c('steelblue', 
             'orange'
             )

# Top 10 Phyla
Phylum.summary <- aggregate(Abundance~Group + Phylum, 
                            data = ps_ra_melt, 
                            FUN = mean)
Phylum.summary.ordered <- Phylum.summary [order(Phylum.summary $ Abundance, decreasing = T),]

top10_phy <- Phylum.summary.ordered %>%
  group_by(Group) %>%
  pull(Phylum) %>%
  unique() %>%
  head(n = 10)

# fig (Boxplot top10 Phylum)
ps_ra_melt $ Abundance_percent <- ifelse(ps_ra_melt$Abundance < 0.001, 0.1,ps_ra_melt$Abundance*100)

# Boxplot
F1 <- ggplot(ps_ra_melt[ps_ra_melt$Phylum %in% top10_phy,], 
             aes(y = Phylum, x = Abundance_percent)) + 
  geom_boxplot(position =position_dodge2(reverse = TRUE), aes(color = Group), alpha = 0.5) + 
  geom_boxplot(position =position_dodge2(reverse = TRUE), aes(fill = Group), alpha = 0.5, outlier.shape = 21) +
  facet_wrap(~ Timepoint, nrow = 1) +
  scale_fill_manual(values = palette) + 
  scale_color_manual(values = palette) + 
  scale_x_log10() +
  ylab("") +
  xlab("Abundance (%)") +
  theme_bw() + theme(axis.title = element_text(size = 10), axis.text.x = element_text(size = 8),legend.position = 'bottom',axis.text.y = element_text(face="italic",size = 8),
  plot.margin = unit(c(1, 1, 1, 1), "lines")) # Adjust the image margins here (top, right, bottom, left)

print(F1)

# Legend
g_legend<-function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)} 
mytheme <- theme(axis.title = element_text(size = 10), axis.text = element_text(size = 8)) + theme_bw()

# plot legend
F1_legend <- g_legend(F1 + geom_boxplot(aes(color=Group,fill=Group),alpha=0.5)+ 
  theme_bw() + theme(axis.title = element_text(size = 10), 
                     axis.text = element_text(size = 8), 
                     axis.text.x=element_blank(), 
                     axis.ticks.x=element_blank(), 
                     legend.position = 'bottom'))

# Save 
FileName <- paste(over_Group_Timepoint,"/Top Phylum Abundance by group and timepoint.pdf", sep = "")
ggsave(FileName, plot = F1, width = 180, height = 170, units = "mm", dpi = 300)

FileName <- paste(over_Group_Timepoint,"/Top Phylum Abundance by group and timepoint.png", sep = "")
ggsave(FileName, plot = F1, width = 180, height = 170, units = "mm", dpi = 300)

# Export to excel
FileName <- paste(over_Group_Timepoint,"/Top Phylum Abundance by group and timepoint.xlsx", sep = "")
write_xlsx(ps_ra_melt, FileName)

# clean environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

## 4.2. Top 10 Family - Group1 x Group2 composition by timepoint
```{r family_abundance_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Top 10 family abundance by group'}

# path
path <- here::here()
over_Group_Timepoint = paste(path,"/overview_by_group_and_timepoint/",sep = "")

# Load aglomerate data by family 
ps <- readRDS("ps1.dna.family.rds")

# Relative abundance values
ps_ra <- transform_sample_counts(ps, function(x) x / sum(x))
ps_ra_melt <- psmelt(ps_ra)

# Color palette
palette <- c("steelblue", 
             "orange")

# # Top 10 Families 
family.summary <- aggregate(Abundance~Group+Family, data = ps_ra_melt, FUN = mean)
family.summary.ordered <- family.summary[order(family.summary$Abundance,decreasing = T),]

top10_fam <- family.summary.ordered %>%
  group_by(Group) %>%
  pull(Family) %>%
  unique() %>%
  head(n = 10)

# Boxplot top family por Timepoint
ps_ra_melt$Abundance_percent <- ifelse(ps_ra_melt$Abundance < 0.001, 0.1,ps_ra_melt$Abundance*100)

F2 <- ggplot(ps_ra_melt[ps_ra_melt$Family %in% top10_fam,], aes(y = Family, x = Abundance_percent)) + 
  geom_boxplot(position =position_dodge2(reverse = TRUE), aes(color = Group), alpha = 0.5) + 
  geom_boxplot(position =position_dodge2(reverse = TRUE), aes(fill = Group), alpha = 0.5, outlier.shape = 21) +
  facet_wrap(~ Timepoint, nrow = 1) +
  scale_fill_manual(values = palette) + 
  scale_color_manual(values = palette) + 
  scale_x_log10() +
  ylab("") +
  xlab("Abundance (%)") +
  theme_bw() + theme(axis.title = element_text(size = 10), axis.text.x = element_text(size = 8),legend.position = 'bottom',axis.text.y = element_text(face="italic",size = 8),
  plot.margin = unit(c(1, 1, 1, 1), "lines")) # Adjust the image margins here (top, right, bottom, left)

print(F2)

# Legend
g_legend<-function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)} 
mytheme <- theme(axis.title = element_text(size = 10), axis.text = element_text(size = 8)) + theme_bw()

# plot legend
F2_legend <- g_legend(F2 + geom_boxplot(aes(color=Group,fill=Group),alpha=0.5)+ 
  theme_bw() + theme(axis.title = element_text(size = 10), 
                     axis.text = element_text(size = 8), 
                     axis.text.x=element_blank(), 
                     axis.ticks.x=element_blank(), 
                     legend.position = 'bottom'))

# save
FileName <- paste(over_Group_Timepoint,"/Top family abundance by group and timepoint.pdf", sep = "")
ggsave(FileName, plot = F2, width = 180, height = 170, units = "mm", dpi = 300)

FileName <- paste(over_Group_Timepoint,"/Top family abundance by group and timepoint.png", sep = "")
ggsave(FileName, plot = F2, width = 180, height = 170, units = "mm", dpi = 300)

# Export to excel
FileName <- paste(over_Group_Timepoint,"/Top family abundance by group and timepoint.xlsx", sep = "")
write_xlsx(ps_ra_melt, FileName)

# clean environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

## 4.3. Top 10 Genus - Group1 x Group2 composition by timepoint 
```{r genus_abundance_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Top 10 genus abundance by group'}

# path
path <- here::here() 
over_Group_Timepoint = paste(path,"/overview_by_group_and_timepoint/",sep = "")

# Load aglomerate data by genus 
ps <- readRDS("ps1.dna.genus.rds")

# Abundance relative values
ps_ra <- transform_sample_counts(ps, function(x) x / sum(x))
ps_ra_melt <- psmelt(ps_ra)

# Color palette
palette <- c("steelblue", 
             "orange")

# Top 10 genus
genus.summary <- aggregate(Abundance~Group+Genus, data = ps_ra_melt, FUN = mean)
genus.summary.ordered <- genus.summary[order(genus.summary$Abundance,decreasing = T),]

top10_genus <- genus.summary.ordered %>%
  group_by(Group) %>%
  pull(Genus) %>%
  unique() %>%
  head(n = 10) 

# boxplot top genera by timepint
ps_ra_melt$Abundance_percent <- ifelse(ps_ra_melt$Abundance < 0.001, 0.1,ps_ra_melt$Abundance*100)

F3 <- ggplot(ps_ra_melt[ps_ra_melt$Genus %in% top10_genus,], aes(y = Genus, x = Abundance_percent)) + 
  geom_boxplot(position =position_dodge2(reverse = TRUE), aes(color = Group), alpha = 0.5) + 
  geom_boxplot(position =position_dodge2(reverse = TRUE), aes(fill = Group), alpha = 0.5, outlier.shape = 21) +
  facet_wrap(~ Timepoint, nrow = 1) +
  scale_fill_manual(values = palette) + 
  scale_color_manual(values = palette) + 
  scale_x_log10() +
  ylab("") +
  xlab("Abundance (%)") +
  theme_bw() + theme(axis.title = element_text(size = 10), axis.text.x = element_text(size = 8),legend.position = 'bottom',axis.text.y = element_text(face="italic",size = 8),
  plot.margin = unit(c(1, 1, 1, 1), "lines")) # Adjust the image margins here (top, right, bottom, left)

print(F3)

# Legend
g_legend<-function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)} 
mytheme <- theme(axis.title = element_text(size = 10), axis.text = element_text(size = 8)) + theme_bw()

# plot legend
F3_legend <- g_legend(F3 + geom_boxplot(aes(color=Group,fill=Group),alpha=0.5)+ 
  theme_bw() + theme(axis.title = element_text(size = 10), 
                     axis.text = element_text(size = 8), 
                     axis.text.x=element_blank(), 
                     axis.ticks.x=element_blank(), 
                     legend.position = 'bottom'))

# save
FileName <- paste(over_Group_Timepoint,"/Top genera gbundance by group and timepoint.pdf", sep = "")
ggsave(FileName, plot = F3, width = 180, height = 170, units = "mm", dpi = 300)

FileName <- paste(over_Group_Timepoint,"/Top genera abundance by group and timepoint.png", sep = "")
ggsave(FileName, plot = F3, width = 180, height = 170, units = "mm", dpi = 300)

# Export to excel
FileName <- paste(over_Group_Timepoint,"/Top genera abundance by group and timepoint.xlsx", sep = "")
write_xlsx(ps_ra_melt, FileName)

# clean environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

## 5.0 Session info
```{r session_info, eval=TRUE, echo=F}

sessionInfo()

```