---
title: 'In this field, insert the project title'
author: 'In this field, insert the author(s) of the analysis'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends 
---

================================================================================ 
# Study description:

~ Include here a brief description of the study ~

================================================================================

# DIFFERENTIAL ABUNDANCE ANALYSIS IN TIME SERIES STUDIES

## 1. Load Libraries
```{r libraries_diffAbund, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

library(gtsummary)
library(writexl)
library(dplyr)
library(phyloseq)
library(nlme)
library(vegan)
library(compositions)
library(ggplot2)
library(lme4)
library(reshape2) 
library(vegan)
library(ade4)
library(plotly)
library(pracma) 
library(fpc) 
library(tidyverse)
library(cluster)
library(RColorBrewer)
library(ape)
library(gplots)
library(RColorBrewer)
library(pheatmap)
library(Maaslin2)
library(ggpicrust2)
library(microbiome)
library(file2meco)
library(microbiomeutilities)
library(gridExtra)
library(GGally)
library(broom)
library(broom.mixed)
library(kableExtra)
library(emmeans)
library(cardx)
library(here)
library(readr)
library(forcats)
library(grid) 
library(combinat)
library(tidyr)
library(tibble)

```

## 2. Differential Abundance Analysis

We performed the differential abundance analysis at genus level with MaAsLin2
```{r dataPrep_maaslin, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here::here()
Dif_abundance = paste(path,"/Differential_abundance_genus/",sep = "")
dir.create(Dif_abundance)

# load phyloseq data and adjust
ps0 <- readRDS("ps1.dna.genus.rds")
ps <- format_to_besthit(ps0)

# preparing the feature table from abundance table
feature_table <- as.data.frame(otu_table(ps))

# extract metadata from ps
metadata <- microbiome::meta(ps)

# creating an interaction variable (Maaslin2 does not accept interaction in modeling)
metadata$Group_Timepoint_interaction <- interaction(metadata$Group, metadata$Timepoint)

# check if the samples are in the columns and transpose the table.
if (taxa_are_rows(ps)) {
  feature_table <- t(feature_table)
}

# clean environment
#rm(list = ls(all = TRUE))

```

```{r run_maaslin, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# run MaAsLin2 
fit_data <- Maaslin2(
    input_data = feature_table,
    input_metadata = metadata,
    output = Dif_abundance,
    fixed_effects = c("Group", "Timepoint", "Group_Timepoint_interaction"),
    random_effects = "Animal",
    reference = c("Group,Group1;Timepoint,TP1;Group_Timepoint_interaction,Group1.TP1"), # Insert here, baseline information of group and timepoint to comparisons
    normalization = "CLR",
    transform = "LOG",      # Logarithmic transformation
    analysis_method = "LM", # Linear model
    min_prevalence = 0.1,
    min_abundance = 0.1,
    correction = "BH",
    standardize = TRUE,
    cores = 8
) #  Maaslin2 output: "significant_results.tsv" file or "fit_data$results"

# save Maaslin2 data
save(fit_data, file = file.path(Dif_abundance, "Maaslin2.RData"))

# clean environment
#rm(list = ls(all = TRUE))

```

## 2.1. Global differential abundance analysis including group, timepoint and group vs timepoint interaction
```{r diffAbund_global, eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=8, fig.height=16}

# paths
path <- here::here()
Dif_abundance <- file.path(path, "Differential_abundance_genus")

load('Maaslin2.RData')

# read significant results
res <- read_tsv(
  file.path(Dif_abundance, "significant_results.tsv"),
  show_col_types = FALSE
)

# baseline levels (reference group) for each variable.
ref_values <- c(
  Group                    = "Group1",
  Group_Timepoint_interaction    = "Group1.TP1",
  Timepoint                = "TP1"
) # Adjust these values according to your analysis (this is the level against which the coefficient was estimated).

# data preparation and creation of comparison labels
df_plot <- res %>%
  filter(metadata != "Group") %>% 
  mutate(
    taxon      = feature,
    coef       = coef,
    direction  = if_else(coef >= 0, "Positive LFC", "Negative LFC"),
    comparison = paste0(
      metadata, ": ", 
      value, 
      " vs ", 
      ref_values[metadata]
    )
  ) %>%
  group_by(comparison) %>%
  slice_max(order_by = abs(coef), n = 10, with_ties = FALSE) %>%
  ungroup()

# horizontal bar plot
p <- ggplot(df_plot, aes(
    y    = reorder(taxon, abs(coef)),
    x    = coef,
    fill = direction
  )) +
  geom_bar(stat = "identity", width = 0.5, color = "black") +
  labs(
    y     = "",
    x     = "Log fold change",
    title = "Differential Abundance (MaAsLin2) - Top genera"
  ) +
  scale_fill_manual(
    values = c(
      "Positive LFC" = "coral", 
      "Negative LFC" = "darkslategray3"
    ),
    name = ""
  ) +
  facet_wrap(~ comparison, scales = "free_y", ncol = 1) +
  theme_bw() +
  theme(
    plot.title       = element_text(hjust = 0),
    axis.text.y      = element_text(face = "italic", size = 8),
    axis.text.x      = element_text(size = 10),
    strip.text.x     = element_text(size = 9, face = "bold"),
    plot.margin      = unit(c(1, 1, 1, 1), "lines"),
    legend.position  = "bottom"
  )

print(p) # display the plot

# Save
ggsave(
  filename = file.path(Dif_abundance, "masalin2_diffabund_comparisons.png"),
  plot     = p,
  width    = 8,
  height   = 16,
  dpi      = 300
)

ggsave(
  filename = file.path(Dif_abundance, "masalin2_diffabund_comparisons.pdf"),
  plot     = p,
  width    = 8,
  height   = 16,
  dpi      = 300
)

```
\FloatBarrier

## 2.2. Visualization of effect size in a Group vs Timepoint Interaction
```{r diffAbund_dataPrep_Vizu, eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=8, fig.height=16}

# paths
path <- here::here()
Dif_abundance <- file.path(path, "Differential_abundance_genus")

# load Maaslin2 object
load("Maaslin2.RData")

# read significant results
res <- read_tsv(
  file.path(Dif_abundance, "significant_results.tsv"),
  show_col_types = FALSE
)
# data for interaction heatmap
interaction_res <- res %>%
  filter(metadata == "Group_Timepoint_interaction") %>%   
  mutate(
    taxon = feature,
    comparison = value,
    LFC = coef
  )

# select taxa with strongest differential effects
top_taxa <- interaction_res %>%
  group_by(taxon) %>%
  summarise(max_effect = max(abs(LFC))) %>%
  arrange(desc(max_effect)) %>%
  slice(1:30) %>%
  pull(taxon)

heatmap_data <- interaction_res %>%
  filter(taxon %in% top_taxa) %>%
  mutate(
    taxon = factor(taxon, levels = rev(unique(taxon)))
  )


```

```{r maaslin2_heatmap_interactionPlot, eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=8, fig.height=16}

# Interaction heatmap plot
p_heatmap <- ggplot(
  heatmap_data,
  aes(
    x = comparison,
    y = taxon,
    fill = LFC
  )
) +
  geom_tile(color = "white") +

  scale_fill_gradient2(
    low = "coral",
    mid = "white",
    high = "darkslategray3",
    midpoint = 0,
    name = "Log fold change"
  ) +

  labs(
    x = "Group-Time comparison",
    y = "",
    title = "MaAsLin2 – Interaction heatmap (Group × Timepoint)"
  ) +

  theme_bw() +

  theme(
    plot.title = element_text(hjust = 0),
    axis.text.y = element_text(
      face = "italic",
      size = 8
    ),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "bottom",
    panel.grid.major = ggplot2::element_blank()
  )

print(p_heatmap) # display heatmap plot

# save heatmap
ggsave(
  filename = file.path(
    Dif_abundance,
    "Maaslin2_interaction_heatmap.png"
  ),
  plot = p_heatmap,
  width = 10,
  height = 8,
  dpi = 300
)

ggsave(
  filename = file.path(
    Dif_abundance,
    "Maaslin2_interaction_heatmap.pdf"
  ),
  plot = p_heatmap,
  width = 10,
  height = 8,
  dpi = 300
)

```

## 2.3. Temporal trajectories at the genus level - clustered visualization
```{r temporalTrajectories_visu, eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=8, fig.height=16}

# paths
path <- here::here()
Dif_abundance <- file.path(path, "Differential_abundance_genus")

# load Maaslin2 object
load(file.path(Dif_abundance, "Maaslin2.RData"))

# identify significant taxa from MaAslin2 results
sig_taxa <- fit_data$results %>%
  dplyr::filter(qval < 0.25) %>%  # MaAslin uses q < 0.25 as default
  dplyr::pull(feature) %>%
  unique()

# extract adjusted values from each sample of the MaAslin2 model
fitted_sig <- fit_data$fitted[rownames(fit_data$fitted) %in% sig_taxa,]
 
# convert fitted matrix to tidy format
fitted_df <- fitted_sig %>%
  as.data.frame() %>%
  tibble::rownames_to_column("taxon") %>%
  tidyr::pivot_longer(
    -taxon,
    names_to = "Sample",
    values_to = "log_abundance"
  )

# add sample metadata
metadata <- microbiome::meta(ps)
metadata$Sample <- rownames(metadata)

fitted_df <- fitted_df %>%
  dplyr::left_join(metadata, by="Sample")

# export df to excel
write_xlsx(fitted_df, file.path(Dif_abundance, "fitted_values.xlsx"))

# calculate mean predicted abundance per condition
predicted <- fitted_df %>%
  dplyr::group_by(taxon, Group, Timepoint) %>%
  dplyr::summarise(
    log_abundance = mean(log_abundance, na.rm = TRUE),
    .groups = "drop"
  )

# convert to taxa × condition matrix
pred_matrix <- predicted %>%
  dplyr::mutate(sample = paste(Group, Timepoint, sep = "_")) %>%
  dplyr::select(taxon, sample, log_abundance) %>%
  tidyr::pivot_wider(
    names_from = sample,
    values_from = log_abundance
  ) %>%
  tibble::column_to_rownames("taxon") %>%
  as.matrix()

# compute row Z-score (standardize taxa profiles)
pred_matrix <- t(scale(t(pred_matrix)))
pred_matrix[is.na(pred_matrix)] <- 0

# hierarchical clustering of taxa
row_clust <- stats::hclust(
  stats::dist(pred_matrix)
)

k <- 3 # Define the number of clusters (according to your data analysis)

clusters <- cutree(row_clust, k = k)

# reorder matrix according to clustering
pred_matrix <- pred_matrix[row_clust$order, ]

# convert clustered matrix to dataframe
cluster_df_plot <- as.data.frame(pred_matrix)

cluster_df_plot$taxon <- rownames(cluster_df_plot)

cluster_df_plot$cluster <- clusters[rownames(pred_matrix)]

cluster_df_plot$taxon <- factor(
  cluster_df_plot$taxon,
  levels = rownames(pred_matrix)
)

# convert matrix to long format
cluster_df_plot <- cluster_df_plot %>%
  tidyr::pivot_longer(
    -c(taxon, cluster),
    names_to = "sample",
    values_to = "zscore"
  ) %>%
  tidyr::separate(
    sample,
    into = c("Group","Timepoint"),
    sep = "_"
  )

# enforce plotting order
cluster_df_plot <- cluster_df_plot %>%
  dplyr::mutate(
    Group = factor(
      Group,
      levels = c("Group1","Group2")
    ),
    Timepoint = factor(
      Timepoint,
      levels = c("TP1","TP2","TP3")
    ),
    cluster = factor(cluster)
  )

# create cluster labels
cluster_df_plot <- cluster_df_plot %>%
  dplyr::mutate(
    cluster_label = factor(
      cluster,
      levels = c(1,2,3,4),
      labels = c(
        "Cluster 1",
        "Cluster 2",
        "Cluster 3",
        "Cluster 4"
        )
    )
  )

```

```{r maaslin2_Test, eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.width=8, fig.height=16}

# clustered longitudinal heatmap plot (using MaAsLin2 fitted values)
p_cluster <- ggplot(
  cluster_df_plot,
  aes(
    x = Timepoint,
    y = taxon,
    fill = zscore
  )
) +
  geom_tile(color = "white") +

  facet_grid(
    cluster_label ~ Group,
    scales = "free_y",
    space = "free_y"
  ) +

  scale_fill_distiller(
  palette = "RdBu",
  direction = -1,
  name = "Row Z-score"
) +

  labs(
    x = "Timepoint",
    y = "Genus",
    title = "Clustered longitudinal heatmap - genera trajectory"
  ) +

  theme_bw() +

  theme(
    axis.text.y = element_text(face = "italic", size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0),
    panel.spacing.x = grid::unit(0, "lines"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


print(p_cluster) # display heatmap plot

# save plot
ggsave(
  filename = file.path(
    Dif_abundance,
    "Maaslin2_longitudinal_clustered_heatmap.png"
  ),
  plot = p_cluster,
  width = 10,
  height = 12,
  dpi = 300
)

ggsave(
  filename = file.path(
    Dif_abundance,
    "Maaslin2_longitudinal_clustered_heatmap.pdf"
  ),
  plot = p_cluster,
  width = 10,
  height = 12,
  dpi = 300
)

```
Genus-level taxa identified by MaAsLin2 were grouped by hierarchical clustering based on their longitudinal Z-score profiles. Clusters represent taxa with shared temporal trajectories across Group1 and Group2 groups.

[The values below should be adjusted according to the analysis]

Cluster 1: Strong increase at D103 after low early abundance 
Cluster 2: Shared decrease toward D103 in PF and DFI
Cluster 3: Evidence of enrichment in the DFI group

### 3. Session info
```{r session_info, eval=TRUE, echo=F}

sessionInfo()

```

