---
title: 'In this field, insert the project title'
author: 'In this field, insert the author(s) of the analysis'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
geometry: "top=2cm, bottom=2cm, left=3cm, right=2cm"
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends
---
================================================================================
# Study description:

~ Include here a brief description of the study ~

Example: Data for this analysis is from the project 'Proteomics and Metagenomics for Identification and Characterization of COVID-19 Biomarkers', ethics approval 4.355.906, Hospital de Clínicas de Porto Alegre (HCPA). The bioinformatics analyses were performed in the Bioinformatics Core of HCPA. This document presents the metagenomics preprocessing workflow for this project.

In this analysis we include 80 nasal and oropharynx swabs from HCPA biobank, collected to perform rt-qPCR for SARS-CoV-2 detection. The samples were selected using COVID-19 severity class (WHO, 2020), as follows: Group 1 (n = 22): positive rt-qPCR for SARS-CoV-2 - COVID-19 - moderate; Group 2 (n = 19, control group): negative rt-qPCR for SARS-CoV-2 (confirmed with a second test), previously classified as moderate COVID-19 by the physician; Group 3 (n = 20): positive rt-qPCR for SARS-CoV-2 - COVID-19 - severe/critical; Group 4 (n = 18, control group): asymptomatic, highly exposed inpatients and healthcare workers, who tested negative by rt-qPCR for SARS-CoV-2. 

================================================================================

### INFORMATION:

The folder that stores the filtered files and those of the other analyses are automatically created in a folder called Pre-processing. Before starting the analysis, it is recommended that:

### RECOMMENDATION:

1 - Create a folder for each project;
2 - Make a copy of the document (.rmd file) to the new project folder you just created;
3 - Link a new project to the created folder (upper right corner of RStudio > New project > Select the already created directory to be the project folder);
4 - If there is more than one sequencing run: to store the .fastq files that will be submitted for analysis, create a new folder for each sequencing run (run1, run2, run3...) within the created project folder.

This workflow is based on the reference https://f1000research.com/articles/5-1492

***IMPORTANT***
If the data originates from different sequencing runs, use the script for BIG DATA.

# PREPROCESSING - SINGLE RUN

## 1. Load libraries
```{r libraries, eval=TRUE, echo=FALSE, warning=FALSE, include = FALSE}

library(knitr)
library(phyloseq)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(dada2)
library(rio)
library(writexl)
library(readxl)
library(dplyr)
library(tidyr)

```

## 2. Setting the paths
```{r paths, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

set.seed(100)

# Define the path to the output files
path <- 'insert/here/the/path'
Preprocessing = paste(path,"/Preprocessing/",sep = "")
dir.create(Preprocessing)

# Defines the paths to the .fastq files 
miseq_path <- file.path('insert/here/the/path') # In this field, you type the path to the folder where the .fastq files are stored
fns <- sort(list.files(miseq_path, full.names = TRUE))
# The commands below will "capture" the identification of the .fastq files. Ensure that the sample names are the same in the files and in the variable that will be linked as the sample identifier (sampleID) in the metadata table. Fastq files with "R1" in the name will be identified as Fwd, and those with "R2" will be identified as Rev. The "name" is what is in front of the first underscore.
fnFs <- fns[grepl("R1", fns)] # Fwd
fnRs <- fns[grepl("R2", fns)] # Rev

save(file = 'NomeDoProjeto_1.RData', list = c('path','Preprocessing','miseq_path','fns','fnFs','fnRs'))

```

## 3. Plot Quality Profile
```{r quality_plot, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=12}

# FWD
plotQualityProfile(fnFs) + ggtitle("Fwd")  
full.fwd.quality.plot <- plotQualityProfile(fnFs) + ggtitle("Fwd") 

#Salvando
FileName <- paste(Preprocessing,"/Fwd Quality Plot.pdf", sep = "")
ggsave(FileName, plot = full.fwd.quality.plot, width = 20, height = 20, units = "in", dpi = 300)

# REV
plotQualityProfile(fnRs) + ggtitle("Rev") 
full.rev.quality.plot <- plotQualityProfile(fnRs) + ggtitle("Rev") 

#Salvando
FileName <- paste(Preprocessing,"/Rev Quality Plot.pdf", sep = "")
ggsave(FileName, plot = full.rev.quality.plot, width = 20, height = 20, units = "in", dpi = 300)

```
\FloatBarrier
In the grayscale, a graph illustrates the frequency of each quality score at every base position. The green line represents the median quality score at each position, while the orange lines depict the quartiles of the quality score distribution. Additionally, the red line indicates the scaled proportion of reads that extend to, at least, that particular position.

## 4. Filter and trim 
```{r filter_and_trim, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# This step involves removing the primers (trimLeft), truncating the sequences (truncLen), and filtering.

load('NomeDoProjeto_1.RData') # Load the previously saved objects here (item 2. Setting the paths) to continue the analysis.

# Define the paths to the filtered files
filt_path <- file.path(Preprocessing, "filtered")
if (!file_test("-d", filt_path)) {
   dir.create(filt_path)
}

filtFs <- file.path(filt_path, basename(fnFs))
filtRs <- file.path(filt_path, basename(fnRs))

# TrimLeft: Size of front and rear primers for removal. 16S v3v4 is 17, 21
# TruncLen: Size to truncate the sequences, based on the plot quality. Try to keep the average quality (green line) above 30 and the sum of both values (fwd, rev) should be at least 480.

out <- data.frame(reads.in = c(), reads.out = c(), sample = c())
for(i in seq_along(fnFs)) {
  z <- fastqPairedFilter(c(fnFs[[i]], fnRs[[i]]),
                    c(filtFs[[i]], filtRs[[i]]),
                    trimLeft=c(17, 21), 
                    truncLen=c(280, 200), truncQ = 6,
                    maxN=0, maxEE=c(2,2), rm.phix = TRUE,
                    compress=TRUE, multithread=12)
  out <- rbind(out, z)
} 

save(file = "NomeDoProjeto_2.RData", list = c('path','Preprocessing','miseq_path','fns','fnFs', 'fnRs','filt_path', 'filtFs', 'filtRs','out'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 5. Dereplication 
```{r derep, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# This step performs sequence dereplication to optimize computational capacity and extracts the sample name from the fastq file names. The name will be everything before the first underscore in the file name and must be identical to the sample identifier used in the metadata table (sampleID). In the metadata table, we must have one identifier per sample, not per file.

load('NomeDoProjeto_2.RData') # Load the previously saved objects here (item 4. Filter and trim) to continue the analysis.

derepFs <- derepFastq(filtFs)
derepRs <- derepFastq(filtRs)

sam_names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1) # The sample identifier will be the name that comes before the first underscore.

names(derepFs) <- sam_names
names(derepRs) <- sam_names

save(file = 'NomeDoProjeto_3.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs', 'out','sam_names', 'derepFs', 'derepRs'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 6. Learning errors
```{r learn_err, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# This step will learn the sequencing error rates, which will be used for later sequence inference.

load('NomeDoProjeto_3.RData')  # Load the previously saved objects here (item 5. Dereplication) to continue the analysis.

ddF <- dada(derepFs, err=NULL, selfConsist=TRUE, MAX_CONSIST = 20, nbases=1e8, multithread = TRUE) 
ddR <- dada(derepRs, err=NULL, selfConsist=TRUE, MAX_CONSIST = 20, nbases=1e8, multithread = TRUE) 

save(file = 'NomeDoProjeto_4.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs','out','sam_names', 'derepFs', 'derepRs', 'ddF', 'ddR'))

```

## 7. Estimating error rates
```{r plot_err_rate, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

load('NomeDoProjeto_4.RData') # Load the previously saved objects here (item 5. Learning errors) to continue the analysis.

# Defines the paths for the output files.
p_errors = paste(Preprocessing,"/plot_errors/",sep = "")
dir.create(p_errors)

# fwd
plotErrorsF <- plotErrors(ddF)
# rev
plotErrorsR <- plotErrors(ddR)

# saving the forward and reverse errors plots
ggsave(plot = plotErrorsF, filename = file.path(p_errors, "plotErrorsF.pdf"), device = "pdf", width = 10, height = 8, units = "in")
ggsave(plot = plotErrorsR, filename = file.path(p_errors, "plotErrorsR.pdf"), device = "pdf", width = 10, height = 8, units = "in")

```

## 8. Sample Inference and merge paired reads
```{r sam_infer, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
#- This step runs the 'core' analysis algorithm, which will infer whether the sequences are true or false. It will extract the sample name from the fastq file names. The identifier will be what comes before the first underscore in the file name and must be identical to the sample identifier used in the metadata table (sampleID). In the metadata table, we must have one identifier per sample, not per file. Also, the sequences will be "assembled" by joining the fwd and rev tapes. For this, a complementarity of at least 12 bases is necessary. If a large part of the data is lost in this step, it means that truncLen (4. Filter and trim) is being too radical. The sum of the truncLen of the fwd and rev tapes must be at least 480.

load('NomeDoProjeto_4.RData') # Load the previously saved objects here (item 5. Learning errors) to continue the analysis.

dadaFs <- dada(derepFs, err=ddF, pool=TRUE, multithread = TRUE) #Pool = TRUE - maior sensibilidade
dadaRs <- dada(derepRs, err=ddR, pool=TRUE, multithread = TRUE)

mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose = TRUE)

save(file = 'NomeDoProjeto_5.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs', 'out','sam_names', 'derepFs', 'derepRs', 'ddF', 'ddR', 'dadaFs', 'dadaRs', 'mergers'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 9. Construct Sequence Table, merge runs, remove chimeras and check the output 
```{r seqtaball_remchim, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# This chunks involves constructing the sequence table, grouping the tables from different runs, and removing chimeras

load('NomeDoProjeto_5.RData') # Load the previously saved objects here (item: 8. Sample Inference and merge paired reads) to continue the analysis.

# Groups all tables
seqtab.all <- makeSequenceTable(mergers[!grepl("Mock", names(mergers))]) # "Mock" is the name of the "mock community - internal control." This line removes the mock from the table.
fileName <- paste(Preprocessing,"/seqtab.all.rds", sep = "")
saveRDS(seqtab.all, fileName)

#dim(seqtab.all) # display df size

# Remove chimeras from the grouped table.
seqtab.nochim <- removeBimeraDenovo(seqtab.all, multithread=TRUE, verbose = TRUE)
fileName <- paste(Preprocessing,"/seqtab.nochim.rds", sep = "")
saveRDS(seqtab.nochim, fileName)

# Checking the percentage of remaining sequences
invisible(capture.output({
  remain_seq <- sum(seqtab.nochim)/sum(seqtab.all)}))
print(paste0(round(remain_seq, 4) * 100, '% remaining sequences.'))

save(file = 'NomeDoProjeto_6.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs','out', 'sam_names', 'derepFs', 'derepRs', 'ddF', 'ddR', 'dadaFs', 'dadaRs', 'mergers', 'seqtab.all', 'seqtab.nochim'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 10.  Remove non-target-length sequences  
```{r rem_non_targ, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
#  Here, sequences outside the target size range are removed. The v3v4 region has a wide size distribution. Therefore, it is possible to observe different sequence sizes, in the range of ~250 to ~500. It is advantageous to remove ASVs outside this range.

load('NomeDoProjeto_6.RData') # Load the previously saved objects here (item: 9. Construct Sequence Table) to continue the analysis.

table(nchar(getSequences(seqtab.nochim)))
seqtab <- seqtab.nochim[,nchar(colnames(seqtab.nochim)) %in% seq(250,500)] # If there are any remaining sequences that are significantly outside the expected range, they are possibly "noise or clutter". Adjust the sequence size range to remove them.
table(nchar(getSequences(seqtab)))

########################################################################################################################################################################################

# Update proposal:

# advantage:
# This avoids removing real sequences and only removes extremely short or long sequences, possibly associated with noise.
# It avoids the use of magic numbers and automates the script.

# # 1) Exploring sequence sizes
#
# tam <- nchar(getSequences(seqtab.nochim))
#
# summary(tam) # display
# table(tam) # display
# hist(tam, breaks = 50,
#      main = "Distribution of ASV size",
#      xlab = " (bp)")
#
#
# # 2) captures the "real" sequences (above 200 bp)
# seqtab.tmp <- seqtab.nochim[, tam >= 200]
#
# # 3) Check the length after the initial filter.
# tam.tmp <- nchar(getSequences(seqtab.tmp))
#
# # 4) Filter based on percentages - defined based on the 1% and 99% of the observed distribution (ADJUSTABLE)
# lim_inf <- quantile(tam.tmp, 0.01)  # 1% Em Ion Torrent: espera-se limites na faixa ~ 220–550 bp
# lim_sup <- quantile(tam.tmp, 0.99)  # 99%
#
# lim_inf # display
# lim_sup # display
#
# # 5) Remove extreme outliers
# seqtab <- seqtab.tmp[, tam.tmp >= lim_inf & tam.tmp <= lim_sup]
#
# # 6) Check the result
# table(nchar(getSequences(seqtab)))

save(file = 'NomeDoProjeto_7.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs','out', 'sam_names', 'derepFs', 'derepRs', 'ddF', 'ddR', 'dadaFs', 'dadaRs', 'mergers', 'seqtab.all', 'seqtab.nochim', 'seqtab'))

```

## 11. Quality control - Tracking of reads during the pre-processing filtering
```{r track_reads, eval=TRUE, echo=TRUE}
# This chunk generates the trace-ability table of the sequences throughout the process.

load('NomeDoProjeto_7.RData') # Load the previously saved objects here (item: 10. Remove non-target-length sequences) to continue the analysis.

# Creates a read tracking table.
getN <- function(x) sum(getUniques(x))

# Combines tracking tables
track.full <- as.data.frame(cbind(out, sapply(dadaFs, getN),sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab)))

# To execute the code below, the samples must be in the same order as the track.full object.

# Importing manually verified and curated data in CSV format
samdf <- read.csv('insert/here/the/path') # In this field, insert the path to the sample dataframe (samdf)  

# Using the sampleID column from the samdf (metadata table) for the track.full object
track.full <- cbind(samdf$Sample_ID, track.full) 

# Set names of columns in dataframe
colnames(track.full) <- c("sampleID","input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")

# Exporting to an Excel file
Preprocessing = paste(path,"/Preprocessing/",sep = "")
FileName <- paste(Preprocessing,"/Reads through the pipeline.xlsx", sep = "")
write_xlsx(track.full,FileName) 

# Adjust formatting to print
trackFull_print <- track.full %>%
  rownames_to_column(var = "old_names") %>%  # transform rownames in column
  column_to_rownames("sampleID") %>%        
  select(-old_names)                        

# table
trackFull_print %>%
  rownames_to_column(var = "sampleID") %>%
  kable(
    caption = "Reads through the pipeline",
    digits  = 0,
    booktabs= TRUE,
    row.names = FALSE
  ) %>%
  kable_classic(full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:5, width = "1.5cm")

```

## 12. Assign taxonomy annotation
```{r assign_taxonomy, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# For taxonomic annotation, you must select the desired database (usually SILVA is the most used, eHOMD for oral and respiratory) and verify that the version is the most up-to-date. Preferably use the most current version. For this, the database file must be prepared (trained) to work with dada2. These trained files are usually available on GitHub or Zenodo, and links can be found in the dada2 or Qiime2 tutorials. This trained file must be downloaded, and the 'path' to this file must be placed in the code below. The following is a link to the trained and certified database files by the creator of dada2. (https://benjjneb.github.io/dada2/training.html).

load('NomeDoProjeto_7.RData') # Load the previously saved objects here (item: 10. Remove non-target-length sequences) to continue the analysis.

# Assign taxonomy
taxtab <- assignTaxonomy(seqtab,'insert/here/the/path', multithread = TRUE) # In this field, insert the path where the trained database is located (*/silva_nr99_v138.2_toGenus_trainset.fa.gz)

# Exporting the table as CSV
fileName <- paste(Preprocessing,"/tabela_sequencias_taxo.csv", sep = "")
write.csv(taxtab, fileName, row.names = FALSE)

save(file = 'NomeDoProjeto_8.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs','out','sam_names', 'derepFs', 'derepRs', 'ddF', 'ddR', 'dadaFs', 'dadaRs', 'mergers', 'seqtab.all', 'seqtab.nochim', 'seqtab', 'taxtab'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 13. Metadata engineering 
```{r metadata_eng, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# if any data engineering is needed before assembling the phyloseq object. This example chunk was used to adjust the HCPA query and merge it with the original metadata table. Data engineering can be very laborious and case-specific. DO NOT RELY ON THE FREE VERSION OF CHATGPT OR ANY OTHER AI FOR THIS TASK (commented on 11/16/2023). If necessary, seek someone with data engineering skills

# Reading the Excel file
curated_dataset <- read_excel("caminho ou nome do arquivo/curated_dataset.xlsx")

# Date/time adjustment
curated_dataset$Extraction_date <- as.POSIXct(curated_dataset$Extraction_date, format = "%Y-%m-%dT%H:%M:%S.%OS", tz = "UTC")

# Grouping the data by "BiobankID" and "Analyte", and creating a new column ranking by extraction date.
curated_dataset <- curated_dataset %>% 
  group_by(BiobankID, Analyte) %>% 
  mutate(rank = dense_rank(Extraction_date))

# Creates a new column of the ranked analyte.
curated_dataset$Analyte_ranked <- paste0(curated_dataset$Analyte, " (", curated_dataset$rank, ")")

# "pivot" of the data based on the ranking of the analytes and the "Result" column.
curated_dataset_wide <- spread(curated_dataset, key = "Analyte_ranked", value = "Result")

# Reading the second Excel file
Lovison_etal_metadata <- read_excel("caminho ou nome para o arquivo_metadata.xlsx")

# Converting "BiobankID" to 'double'
Lovison_etal_metadata$BiobankID <- as.double(Lovison_etal_metadata$BiobankID)

# Joining the two dataframes by "BiobankID"
merged_data <- left_join(Lovison_etal_metadata, curated_dataset_wide, by = "BiobankID")

# Removing unnecessary columns
merged_data <- select(merged_data, -Analyte, -rank, -Extraction_date, -CollectionDate, -Exam, -HospitalLocation)

# Summarize
summarized_data <- merged_data %>%
  group_by(sampleID) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")

summarized_data <- inner_join(Lovison_etal_metadata, summarized_data, by = "sampleID")

# Removing samples that did not meet the quality criteria to "match" the pre-processed data
summarized_data_filtered <- summarized_data %>%
  filter(sampleID != "M02", sampleID != "M09", sampleID != "M49", sampleID != "M71", sampleID != "M79")

# Exporting the processed data as a CSV file
write.csv(summarized_data_filtered, file = "summarized_data_filtered.csv", sep=",", dec=".", row.names = FALSE)

```

## 14. Importing the completed sample dataframe  
```{r import_samdf, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# The rules for creating the metadata table are given to the researchers prior to the analysis. Do not manually adjust tables, as the chances of error are enormous and, if errors occur, they become your responsibility. If the table does not conform to the guidelines, return it to the researcher and request the necessary adjustments. The metadata table is 'mandatory' for the continuation of the analysis, as it provides the variables used in subsequent analyses.

load('NomeDoProjeto_8.RData') # Load the previously saved objects here (item: 12. Assign taxonomy) to continue the analysis.

# Importing sample dataframe in CSV format.
samdf <- read.csv('insert/here/the/path') # In this field, insert the path to the sample dataframe ".csv".

rownames(samdf) <- samdf$Sample_ID # Here, the sample dataframe identifier (sampleID) is linked to the rownames, which is the sample name extracted from the file names in the previous steps. It is crucial that the names are the same, otherwise an error will occur. SampleID is the name of the column (header) that contains the sample identifier. This step is important because several functions use rownames internally in their execution.  

save(file = 'NomeDoProjeto_9.RData', list = c('path','Preprocessing','miseq_path','filt_path','fns','fnFs', 'fnRs', 'filtFs', 'filtRs','out','sam_names', 'derepFs', 'derepRs', 'ddF', 'ddR', 'dadaFs', 'dadaRs', 'mergers', 'seqtab.all', 'seqtab.nochim', 'seqtab', 'taxtab','samdf'))

```

## 15. Construct the phyloseq object
```{r phyloseq_object, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

load('NomeDoProjeto_9.RData') # Load the previously saved objects here (item: 14. Importing the ready metadata table) to continue the analysis.

# Producing the phyloseq object
ps <- phyloseq(tax_table(taxtab),
               sample_data(samdf),
               otu_table(seqtab, taxa_are_rows = FALSE))
ps
saveRDS(ps, file = "ps.rds")  # Saving the raw ps object. In this object, the sequence name will be the sequence itself. 

# Renaming the sequences for better tabulation
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps.dna <- merge_phyloseq(ps, dna)
taxa_names(ps.dna) <- paste0("ASV", seq(ntaxa(ps.dna)))
ps.dna

saveRDS(ps.dna, file = "ps.dna.rds") # In this object, the sequence name will be ASV and a corresponding number. This object will be the usual one for the analysis. Throughout the analysis, for figures where it is necessary to format the names with the taxonomy, the 'format_to_best_hit' function should be used, which will possibly already be set in the chunk. This formatting cannot be done beforehand because some analysis functions do not tolerate this formatting.  

ps <- readRDS("ps.rds")
ps.dna <- readRDS("ps.dna.rds")

# Saving only the phyloseq objects
save(file = 'NomeDoProjeto_10.RData', list = c('ps', 'ps.dna'))

# Generating abundance table + taxonomy in CSV file

# Extracting abundance table
otu <- as.data.frame(otu_table(ps))
# Ensuring that ASVs are on the lines
if (!taxa_are_rows(ps)) {
  otu <- t(otu)
}
# Extract taxonomic table
tax <- as.data.frame(tax_table(ps))
# Combine abundance and taxonomy data
ASV_Table <- cbind(tax, otu)
# Rename the ASV column (Transform row names (ASV) into a column)
ASV_Table <- cbind(ASV = rownames(ASV_Table), ASV_Table)
rownames(ASV_Table) <- NULL
# Generates the CSV file
write.csv(ASV_Table, file = file.path(Preprocessing,"ASV_Table.csv"), row.names = FALSE, quote = FALSE)

```

## 16. Session info
```{r session_info, eval=TRUE, echo=FALSE}

sessionInfo()

```
**NomeDoProjeto_10.RData** Record: this object contains the output of phyloseq for further analysis.
