---
title: 'In this field, insert the project title'
author: 'In this field, insert the author(s) of the analysis'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
geometry: "top=2cm, bottom=2cm, left=3cm, right=2cm"
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends
---
================================================================================
# Study description:

~ Include here a brief description of the study ~


================================================================================

INFORMATION:

ALL COMPUTATIONALLY HEAVY/TIME-CONSUMING STEPS INCLUDE A SAVE FOR PROCESS SAFETY.

This script was developed to analyze 3 different races simultaneously and needs to be adapted if this number changes.

The folder that stores the filtered files and those of the other analyses are automatically created in a folder called Pre-processing. Before starting the analysis, it is recommended that:

RECOMMENDATION:

1 - Create a folder for each project;
2 - Make a copy of the document (.rmd file) to the new project folder you just created;
3 - Link a new project to the created folder (upper right corner of RStudio > New project > Select the already created directory to be the project folder);
4 - If there is more than one sequencing run: to store the .fastq files that will be submitted for analysis, create a new folder for each sequencing run (run1, run2, run3...) within the created project folder.


# PREPROCESSING BIG DATA

## 1. Load libraries
```{r libraries, eval=TRUE, echo=FALSE, warning=FALSE, include = FALSE}

library(knitr)
library(phyloseq)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(dada2)
library(rio)
library(writexl)
library(readxl)
library(dplyr)
library(tidyr)
library(here)

```

## 2. Setting the paths
```{r paths, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

set.seed(100)

# Define the path to the output files
path <- here::here()

# Define the paths to the fastq files for the different sequencing runs
path1 <- 'insert/here/the/path'
path2 <- 'insert/here/the/path'
path3 <- 'insert/here/the/path'

# Defines the paths to the processing (output) files
Preprocessing = paste(path,"/Preprocessing/",sep = "")
dir.create(Preprocessing)

# The lines below will 'capture' the identification of the .fastq files. Ensure that the sample names are the same in the files and in the variable that will be linked as the sample identifier (Sample_ID) in the metadata table. Fastq files with "R1" in the name will be identified as Fwd. The "name" is what is in front of the first underscore.

# List .fastq files for each run
fnFs1 <- sort(list.files(path1, pattern="_R1", full.names = TRUE))
fnRs1 <- sort(list.files(path1, pattern="_R2", full.names = TRUE))
fnFs2 <- sort(list.files(path2, pattern="_R1", full.names = TRUE))
fnRs2 <- sort(list.files(path2, pattern="_R2", full.names = TRUE))
fnFs3 <- sort(list.files(path3, pattern="_R1", full.names = TRUE))
fnRs3 <- sort(list.files(path3, pattern="_R2", full.names = TRUE))

save(file = 'NomeDoProjeto_1.RData', list = c('path','path1','path2','path3','Preprocessing','fnFs1', 'fnRs1', 'fnFs2','fnRs2', 'fnFs3', 'fnRs3'))

```

## 3. Plot Quality Profile
```{r quality_plot1F, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="0.8\\linewidth", fig.cap="Quality profile of forward reads (run 1)"}
# View the quality of the files for each run.

# The quality doesn't usually vary much between samples from the same run, but if you are analyzing a large number of samples, 6 samples is considered a good number for visualization and should be enclosed in square brackets, example: (fnFs1[1:6]).

# First sequencing run

# FWD
plotQualityProfile(fnFs1[1:5]) + ggtitle("Fwd1")
full.fwd.quality.plot1 <- plotQualityProfile(fnFs1[1:5]) + ggtitle("Fwd") 

# saving the plots
FileName <- paste(Preprocessing,"/Fwd Quality Plot1.pdf", sep = "")
ggsave(FileName, plot = full.fwd.quality.plot1, width = 20, height = 20, units = "in", dpi = 300)

```

```{r quality_plot1R, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="0.8\\linewidth", fig.cap="Quality profile of reverse reads (run 1)"}
# First sequencing run

# REV
plotQualityProfile(fnRs1[1:5]) + ggtitle("Rev1")
full.rev.quality.plot1 <- plotQualityProfile(fnRs1[1:5]) + ggtitle("Rev") 
# saving the plots
FileName <- paste(Preprocessing,"/Rev Quality Plot1.pdf", sep = "")
ggsave(FileName, plot = full.rev.quality.plot1, width = 20, height = 20, units = "in", dpi = 300)

```

```{r quality_plot2F, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="0.8\\linewidth", fig.cap="Quality profile of forward reads (run 2)"}
# Second sequencing run 

# FWD
plotQualityProfile(fnFs2[1:5]) + ggtitle("Fwd2")
full.fwd.quality.plot2 <- plotQualityProfile(fnFs2[1:5]) + ggtitle("Fwd") 
# saving the plots
FileName <- paste(Preprocessing,"/Fwd Quality Plot2.pdf", sep = "")
ggsave(FileName, plot = full.fwd.quality.plot2, width = 20, height = 20, units = "in", dpi = 300)

```

```{r quality_plot2R, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="0.8\\linewidth", fig.cap="Quality profile of reverse reads (run 2)"}
# Second sequencing run

# REV
plotQualityProfile(fnRs2[1:5]) + ggtitle("Rev2")
full.rev.quality.plot2 <- plotQualityProfile(fnRs2[1:5]) + ggtitle("Rev") 
# saving the plots
FileName <- paste(Preprocessing,"/Rev Quality Plot2.pdf", sep = "")
ggsave(FileName, plot = full.rev.quality.plot2, width = 20, height = 20, units = "in", dpi = 300)

```

```{r quality_plot3F, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="0.8\\linewidth", fig.cap="Quality profile of forward reads (run 3)"}
# Third sequencing run

# FWD
plotQualityProfile(fnFs3[1:5]) + ggtitle("Fwd3")
full.fwd.quality.plot3 <- plotQualityProfile(fnFs3[1:5]) + ggtitle("Fwd") 
# saving the plots
FileName <- paste(Preprocessing,"/Fwd Quality Plot3.pdf", sep = "")
ggsave(FileName, plot = full.fwd.quality.plot3, width = 20, height = 20, units = "in", dpi = 300)

```

```{r quality_plot3R, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=6, out.width="0.8\\linewidth", fig.cap="Quality profile of reverse reads (run 3)"}
# Third sequencing run

# REV
plotQualityProfile(fnRs3[1:5]) + ggtitle("Rev3")
full.rev.quality.plot3 <- plotQualityProfile(fnRs3[1:5]) + ggtitle("Rev") 
# saving the plots
FileName <- paste(Preprocessing,"/Rev Quality Plot3.pdf", sep = "")
ggsave(FileName, plot = full.rev.quality.plot3, width = 20, height = 20, units = "in", dpi = 300)

```
\FloatBarrier

In the grayscale, a graph illustrates the frequency of each quality score at every base position. The green line represents the median quality score at each position, while the orange lines depict the quartiles of the quality score distribution. Additionally, the red line indicates the scaled proportion of reads that extend to, at least, that particular position.

## 4. Filter and trim
```{r filter_and_trim, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# This step performs primer removal (trimLeft), sequence truncation (truncLen), and filtering.

load('NomeDoProjeto_1.RData') # Load the previously saved objects here (item 2. Setting the paths) to continue the analysis.

# Define paths for the filtered files
filt_path1 <- file.path(path1, "filtered")
filt_path2 <- file.path(path2, "filtered")
filt_path3 <- file.path(path3, "filtered")

filtFs1 <- file.path(filt_path1, basename(fnFs1))
filtRs1 <- file.path(filt_path1, basename(fnRs1))
filtFs2 <- file.path(filt_path2, basename(fnFs2))
filtRs2 <- file.path(filt_path2, basename(fnRs2))
filtFs3 <- file.path(filt_path3, basename(fnFs3))
filtRs3 <- file.path(filt_path3, basename(fnRs3))

# Filtering and trimming for each run

# Main parameters of the dada2:
# TrimLeft: Size of the front and rear primers for removal. 16S v3v4 is 17, 21.
# TruncLen: Size to truncate the sequences, based on the quality plot. Try to keep the average quality (green line) above 30 and the sum of both values (front and rear) should be at least 480.

out1 <- filterAndTrim(fnFs1, filtFs1, fnRs1, filtRs1, trimLeft=c(17, 21), truncLen=c(280, 200), maxN=0, maxEE=c(2,2), truncQ=6, rm.phix=TRUE, compress=TRUE, multithread=12)
out2 <- filterAndTrim(fnFs2, filtFs2, fnRs2, filtRs2, trimLeft=c(17, 21), truncLen=c(280, 200), maxN=0, maxEE=c(2,2), truncQ=6, rm.phix=TRUE, compress=TRUE, multithread=12)
out3 <- filterAndTrim(fnFs3, filtFs3, fnRs3, filtRs3, trimLeft=c(17, 21), truncLen=c(280, 200), maxN=0, maxEE=c(2,2), truncQ=6, rm.phix=TRUE, compress=TRUE, multithread=12)

save(file = "NomeDoProjeto_2.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'filtFs1', 'filtRs1', 'filtFs2', 'filtRs2', 'filtFs3', 'filtRs3'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 5. Learning errors 
```{r learn_err, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

# This step will learn the sequencing error rates, which will be used for later sequence inference.

load('NomeDoProjeto_2.RData') # Load the previously saved objects here (4. Filter and trim) to continue the analysis.

# File preparation (renames the files)
sample.names1F <- sapply(strsplit(basename(filtFs1), "_"), `[`, 1) # Assumes filename = samplename_XXX.fastq.gz
sample.names1R <- sapply(strsplit(basename(filtRs1), "_"), `[`, 1) # Assumes filename = samplename_XXX.fastq.gz
sample.names2F <- sapply(strsplit(basename(filtFs2), "_"), `[`, 1) # Assumes filename = samplename_XXX.fastq.gz
sample.names2R <- sapply(strsplit(basename(filtRs2), "_"), `[`, 1) # Assumes filename = samplename_XXX.fastq.gz
sample.names3F <- sapply(strsplit(basename(filtFs3), "_"), `[`, 1) # Assumes filename = samplename_XXX.fastq.gz
sample.names3R <- sapply(strsplit(basename(filtRs3), "_"), `[`, 1) # Assumes filename = samplename_XXX.fastq.gz

if(!identical(sample.names1F, sample.names1R)) stop("Forward and reverse files do not match.")
if(!identical(sample.names2F, sample.names2R)) stop("Forward and reverse files do not match.")
if(!identical(sample.names3F, sample.names3R)) stop("Forward and reverse files do not match.")

names(filtFs1) <- sample.names1F
names(filtRs1) <- sample.names1F
names(filtFs2) <- sample.names2F
names(filtRs2) <- sample.names2F
names(filtFs3) <- sample.names3F
names(filtRs3) <- sample.names3F

set.seed(100)

# Learn forward error rates
errF1 <- learnErrors(filtFs1, nbases=1e8, multithread=12)
errF2 <- learnErrors(filtFs2, nbases=1e8, multithread=12)
errF3 <- learnErrors(filtFs3, nbases=1e8, multithread=12)

# Learn reverse error rates
errR1 <- learnErrors(filtRs1, nbases=1e8, multithread=12)
errR2 <- learnErrors(filtRs2, nbases=1e8, multithread=12)
errR3 <- learnErrors(filtRs3, nbases=1e8, multithread=12)

save(file = "NomeDoProjeto_3.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'sample.names1F', 'sample.names1R', 'filtFs1', 'filtRs1', 'sample.names2F', 'sample.names2R','filtFs2', 'filtRs2', 'sample.names3F', 'sample.names3R','filtFs3','filtRs3','errF1', 'errF2', 'errF3','errR1', 'errR2', 'errR3'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line

```

## 6. Estimating error rates
```{r plot_err_rate, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

load('NomeDoProjeto_3.RData') # Load the previously saved objects here (item 5. Learning errors) to continue the analysis.

# Defines the paths to the output files
p_errors = paste(Preprocessing,"/plot_errors/",sep = "")
dir.create(p_errors)

plotErrorsF1 = plotErrors(errF1, nominalQ=TRUE) + ggtitle("Fwd1")
plotErrorsF2 = plotErrors(errF2, nominalQ=TRUE) + ggtitle("Fwd2")
plotErrorsF3 = plotErrors(errF3, nominalQ=TRUE) + ggtitle("Fwd3")

plotErrorsR1 = plotErrors(errR1, nominalQ=TRUE) + ggtitle("Rev1")
plotErrorsR2 = plotErrors(errR2, nominalQ=TRUE) + ggtitle("Rev2")
plotErrorsR3 = plotErrors(errR3, nominalQ=TRUE) + ggtitle("Rev3")

# Saving the Fs error graphs as PDFs
ggsave(plot = plotErrorsF1, filename = file.path(p_errors, "plotErrorsF1.pdf"), device = "pdf", width = 10, height = 8, units = "in")
ggsave(plot = plotErrorsF2, filename = file.path(p_errors, "plotErrorsF2.pdf"), device = "pdf", width = 10, height = 8, units = "in")
ggsave(plot = plotErrorsF3, filename = file.path(p_errors, "plotErrorsF3.pdf"), device = "pdf", width = 10, height = 8, units = "in")

# Saving the Rs error graphs as PDFs
ggsave(plot = plotErrorsR1, filename = file.path(p_errors, "plotErrorsR1.pdf"), device = "pdf", width = 10, height = 8, units = "in")
ggsave(plot = plotErrorsR2, filename = file.path(p_errors, "plotErrorsR2.pdf"), device = "pdf", width = 10, height = 8, units = "in")
ggsave(plot = plotErrorsR3, filename = file.path(p_errors, "plotErrorsR3.pdf"), device = "pdf", width = 10, height = 8, units = "in")

```

## 7. Sample inference and merger of paired-end reads 
```{r sam_inf_merg, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# This step runs the 'core' analysis algorithm, which infers whether the sequences are true or false and extracts the sample name from the fastq file names. The name will be everything before the first underscore in the file name and must be identical to the sample identifier used in the metadata table (sampleID). In the metadata table, we must have one identifier per sample, not per file. Additionally, it 'assembles' the sequences, joining the fwd and rev. For this, a complementarity of at least 12 databases is necessary. If a large part of the data is lost in this step, it means that the truncLen of filterandtrim (step 0.3.4) was too radical. The sum of the truncLen of the fwd and rev must be at least 480.

load('NomeDoProjeto_3.RData') # Load the previously saved objects here (item 5. Learning errors) to continue the analysis.

mergers1 <- vector("list", length(sample.names1F))
names(mergers1) <- sample.names1F
for(sam in sample.names1F) {
  cat("Processing:", sam, "\n")
    derepF <- derepFastq(filtFs1[[sam]])
    ddF1 <- dada(derepF, err=errF1, multithread=12)
    derepR <- derepFastq(filtRs1[[sam]])
    ddR1 <- dada(derepR, err=errR1, multithread=12)
    merger1 <- mergePairs(ddF1, derepF, ddR1, derepR)
    mergers1[[sam]] <- merger1
}
rm(derepF); rm(derepR)

mergers2 <- vector("list", length(sample.names2F))
names(mergers2) <- sample.names2F
for(sam in sample.names2F) {
  cat("Processing:", sam, "\n")
    derepF <- derepFastq(filtFs2[[sam]])
    ddF2 <- dada(derepF, err=errF2, multithread=12)
    derepR <- derepFastq(filtRs2[[sam]])
    ddR2 <- dada(derepR, err=errR2, multithread=12)
    merger2 <- mergePairs(ddF2, derepF, ddR2, derepR)
    mergers2[[sam]] <- merger2
}
rm(derepF); rm(derepR)

mergers3 <- vector("list", length(sample.names3F))
names(mergers3) <- sample.names3F
for(sam in sample.names3F) {
  cat("Processing:", sam, "\n")
    derepF <- derepFastq(filtFs3[[sam]])
    ddF3 <- dada(derepF, err=errF3, multithread=12)
    derepR <- derepFastq(filtRs3[[sam]])
    ddR3 <- dada(derepR, err=errR3, multithread=12)
    merger3 <- mergePairs(ddF3, derepF, ddR3, derepR)
    mergers3[[sam]] <- merger3
}

save(file = "NomeDoProjeto_4.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'sample.names1F', 'sample.names1R', 'sample.names2F', 'sample.names2R', 'sample.names3F', 'sample.names3R','errF1', 'errF2', 'errF3', 'errR1', 'errR2', 'errR3','mergers1', 'ddF1', 'ddR1', 'mergers2', 'ddF2', 'ddR2', 'mergers3', 'ddF3', 'ddR3'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 8. Construct Sequence Table, merge runs, remove chimeras and check the output 
```{r seqtaball_remchim, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# This chunk involves constructing the sequence table, grouping the tables from different runs, and removing chimeras

load('NomeDoProjeto_4.RData') # Carregue aqui os objetos salvos anteriormente (item 7. Sample inference and merger of paired-end reads) para continuar as analises.

# Construct the sequence tables
seqtab1 <- makeSequenceTable(mergers1[!grepl("Mock", names(mergers1))]) #"Mock" é o nome da "mock community - controle interno. Essa linha remove a mock da tabela
seqtab2 <- makeSequenceTable(mergers2[!grepl("Mock", names(mergers2))])
seqtab3 <- makeSequenceTable(mergers3[!grepl("Mock", names(mergers3))])

# Groups all tables
seqtab.all <- mergeSequenceTables(seqtab1, seqtab2, seqtab3)
fileName <- paste(Preprocessing,"/seqtab.all.rds", sep = "")
saveRDS(seqtab.all, fileName)

# Remove chimeras from the grouped table
seqtab.nochim <- removeBimeraDenovo(seqtab.all, method="consensus", multithread=12, verbose=TRUE)
fileName <- paste(Preprocessing,"/seqtab.nochim.rds", sep = "")
saveRDS(seqtab.nochim, fileName)

# Check the percentage of remaining sequences
sum(seqtab.nochim)/sum(seqtab.all)

save(file = "NomeDoProjeto_5.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'sample.names1F', 'sample.names1R', 'sample.names2F', 'sample.names2R', 'sample.names3F', 'sample.names3R','errF1', 'errF2', 'errF3','errR1', 'errR2', 'errR3', 'mergers1', 'ddF1', 'ddR1', 'mergers2', 'ddF2', 'ddR2', 'mergers3', 'ddF3', 'ddR3', 'seqtab.all', 'seqtab.nochim'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 9. Remove non-target-length sequences 
```{r rem_non_targ, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
#  Here, sequences outside the target size range are removed. The v3v4 region has a wide size distribution. Therefore, it is possible to observe different sequence sizes, in the range of ~250 to ~500. It is advantageous to remove ASVs outside this range.

load('NomeDoProjeto_5.RData') # Load the previously saved objects here (item: 8. Construct Sequence Table) to continue the analysis.

table(nchar(getSequences(seqtab.nochim)))
seqtab <- seqtab.nochim[,nchar(colnames(seqtab.nochim)) %in% seq(250,500)] # If there are any remaining sequences that are significantly outside the expected range, they are possibly "noise or clutter". Adjust the sequence size range to remove them. 
table(nchar(getSequences(seqtab)))

#=========================================================================================================================

# Update proposal (thaianeNascimento):

# In Ion Torrent data, the size of the reads shows greater variability. Wider limits or limits based on the observed distribution were used.
# This avoids removing real sequences and only removes extremely short or long sequences, possibly associated with noise.

# # 1) Exploring sequence sizes
#
# tam <- nchar(getSequences(seqtab.nochim))
#
# summary(tam) # display
# table(tam) # display
# hist(tam, breaks = 50,
#      main = "Distribution of ASV size",
#      xlab = " (bp)")
#
#
# # 2) captures the "real" sequences (above 200 bp)
# seqtab.tmp <- seqtab.nochim[, tam >= 200]
#
# # 3) Check the length after the initial filter.
# tam.tmp <- nchar(getSequences(seqtab.tmp))
#
# # 4) Filter based on percentages - defined based on the 1% and 99% of the observed distribution (ADJUSTABLE)
# lim_inf <- quantile(tam.tmp, 0.01)  # 1% Em Ion Torrent: espera-se limites na faixa ~ 220–550 bp
# lim_sup <- quantile(tam.tmp, 0.99)  # 99%
#
# lim_inf # display
# lim_sup # display
#
# # 5) Remove extreme outliers
# seqtab <- seqtab.tmp[, tam.tmp >= lim_inf & tam.tmp <= lim_sup]
#
# # 6) Check the result
# table(nchar(getSequences(seqtab)))

save(file = "NomeDoProjeto_6.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'sample.names1F', 'sample.names1R', 'sample.names2F', 'sample.names2R', 'sample.names3F', 'sample.names3R','errF1', 'errF2', 'errF3','errR1', 'errR2', 'errR3', 'mergers1', 'ddF1', 'ddR1', 'mergers2', 'ddF2', 'ddR2', 'mergers3', 'ddF3', 'ddR3', 'seqtab.all', 'seqtab.nochim','seqtab'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 10. Quality control - Tracking of reads during the pre-processing filtering 
```{r track_reads, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# This chunk generates the trace-ability table of the sequences throughout the process.

load('NomeDoProjeto_6.RData') # Load the previously saved objects here (item: 9. Remove non-target-length sequences) to continue the analysis.

# Create a read tracking table
getN <- function(x) sum(getUniques(x))

# Combine tracking tables
track <- rbind(out1, out2, out3)

# Combine track with grouped table after chimera removal
track.full <- as.data.frame(cbind(track, rowSums(seqtab.all), rowSums(seqtab)))

# To execute the code below, the samples must be in the same order as the track.full object.

# Importing manually verified and curated data in CSV format
samdf <- read.csv('insert/here/the/path') # In this field, insert the path to the sample dataframe (samdf)

#=========================================================================================================================

# If you need to remove an outlier or need to remove any row from the metadata table - execute the following:

# Remove specific sample by sampleID
#track.full <- track.full[track.full$sampleID != "Sample_20", ]

#==========================================================================================================================

# Using the sampleID column from the samdf (metadata table) for the track.full object.
track.full <- cbind(samdf$Sample_ID, track.full)

# Renaming the columns 
colnames(track.full) <- c("sampleID", "input", "filtered", "merged", "nochim")

# Exporting to an Excel file
FileName <- paste(Preprocessing,"/Reads through the pipeline.xlsx", sep = "")
write_xlsx(track.full, FileName) 

# Adjust formatting to print
trackFull_print <- track.full %>%
  rownames_to_column(var = "old_names") %>%  # transform rownames in column
  column_to_rownames("sampleID") %>%        
  select(-old_names)                        

# Printing table
trackFull_print %>%
  rownames_to_column(var = "sampleID") %>%
  kable(
    caption = "Reads through the pipeline",
    digits  = 0,
    booktabs= TRUE,
    row.names = FALSE
  ) %>%
  kable_classic(full_width = FALSE, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:5, width = "1.5cm")

```

## 11. Assign taxonomy annotation
```{r assign_taxonomy, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# For taxonomic annotation, you must select the desired database (usually SILVA is the most used, eHOMD for oral and respiratory) and verify that the version is the most up-to-date. Preferably use the most current version. For this, the database file must be prepared (trained) to work with dada2. These trained files are usually available on GitHub or Zenodo, and links can be found in the dada2 or Qiime2 tutorials. This trained file must be downloaded, and the 'path' to this file must be placed in the code below. The following is a link to the trained and certified database files by the creator of dada2. (https://benjjneb.github.io/dada2/training.html).

load('NomeDoProjeto_6.RData') # Load the previously saved objects here (item: 9. Remove non-target-length sequences) to continue the analysis.

# Assign taxonomy
taxtab <- assignTaxonomy(seqtab.nochim, 'insert/here/the/path', multithread = TRUE) # In this field, insert the path where the trained database is located (*/silva_nr99_v138.2_toGenus_trainset.fa.gz)

save(file = "NomeDoProjeto_7.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'sample.names1F', 'sample.names1R', 'sample.names2F', 'sample.names2R', 'sample.names3F', 'sample.names3R', 'errF1', 'errF2', 'errF3', 'errR1', 'errR2', 'errR3', 'mergers1', 'ddF1', 'ddR1', 'mergers2', 'ddF2', 'ddR2', 'mergers3', 'ddF3', 'ddR3', 'seqtab.all', 'seqtab.nochim', 'seqtab', 'taxtab'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 12. Metadata engineering
```{r metadata_eng, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# if any data engineering is needed before assembling the phyloseq object. This example chunk was used to adjust the HCPA query and merge it with the original metadata table. Data engineering can be very laborious and case-specific. DO NOT RELY ON THE FREE VERSION OF CHATGPT OR ANY OTHER AI FOR THIS TASK (commented on 11/16/2023). If necessary, seek someone with data engineering skills

# Reading the Excel file
curated_dataset <- read_excel("caminho ou nome do arquivo/curated_dataset.xlsx")

# Date/time adjustment
curated_dataset$Extraction_date <- as.POSIXct(curated_dataset$Extraction_date, format = "%Y-%m-%dT%H:%M:%S.%OS", tz = "UTC")

# Grouping the data by "BiobankID" and "Analyte", and creating a new column ranking by extraction date
curated_dataset <- curated_dataset %>% 
  group_by(BiobankID, Analyte) %>% 
  mutate(rank = dense_rank(Extraction_date))

# Creates a new column of the ranked analyte
curated_dataset$Analyte_ranked <- paste0(curated_dataset$Analyte, " (", curated_dataset$rank, ")")

# "pivot" of the data based on the ranking of the analytes and the "Result" column.
curated_dataset_wide <- spread(curated_dataset, key = "Analyte_ranked", value = "Result")

# Reading the second Excel file
Lovison_etal_metadata <- read_excel("caminho ou nome para o arquivo_metadata.xlsx")

# Converting "BiobankID" to 'double'
Lovison_etal_metadata$BiobankID <- as.double(Lovison_etal_metadata$BiobankID)

# Joining the two dataframes by "BiobankID"
merged_data <- left_join(Lovison_etal_metadata, curated_dataset_wide, by = "BiobankID")

# Removing unnecessary columns
merged_data <- select(merged_data, -Analyte, -rank, -Extraction_date, -CollectionDate, -Exam, -HospitalLocation)

# Summarize
summarized_data <- merged_data %>%
  group_by(sampleID) %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")

summarized_data <- inner_join(Lovison_etal_metadata, summarized_data, by = "sampleID")

# Removing samples that did not meet the quality criteria to "match" the pre-processed data.
summarized_data_filtered <- summarized_data %>%
  filter(sampleID != "M02", sampleID != "M09", sampleID != "M49", sampleID != "M71", sampleID != "M79")

# Exporting the processed data as a CSV file
write.csv(summarized_data_filtered, file = "summarized_data_filtered.csv", sep=",", dec=".", row.names = FALSE)

```

## 13. Importing the completed sample dataframe 
```{r import_samdf, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# The rules for creating the metadata table are given to the researchers prior to the analysis. Do not manually adjust tables, as the chances of error are enormous and, if errors occur, they become your responsibility. If the table does not conform to the guidelines, return it to the researcher and request the necessary adjustments. The metadata table is 'mandatory' for the continuation of the analysis, as it provides the variables used in subsequent analyses.

load('NomeDoProjeto_7.RData') # Load the previously saved objects here (item: 11. Assign taxonomy s) to continue the analysis.

# Importing sample dataframe in CSV format.
samdf <- read.csv('insert/here/the/path') # In this field, insert the path to the sample dataframe ".csv"

rownames(samdf) <- samdf$Sample_ID # Here, the sample dataframe identifier (sampleID) is linked to the rownames, which is the sample name extracted from the file names in the previous steps. It is crucial that the names are the same, otherwise an error will occur. SampleID is the name of the column (header) that contains the sample identifier. This step is important because several functions use rownames internally in their execution. 

save(file = "NomeDoProjeto_8.RData", list = c('path','path1','path2','path3','Preprocessing','out1', 'out2', 'out3', 'sample.names1F', 'sample.names1R', 'sample.names2F', 'sample.names2R', 'sample.names3F', 'sample.names3R', 'errF1', 'errF2', 'errF3', 'errR1', 'errR2', 'errR3', 'mergers1', 'ddF1', 'ddR1', 'mergers2', 'ddF2', 'ddR2', 'mergers3', 'ddF3', 'ddR3', 'seqtab.all', 'seqtab.nochim', 'seqtab', 'taxtab','samdf'))

rm(list = ls()) # Cleaning the environment (to proceed without cleaning the environment, comment "#" this line)

```

## 14. Construct the phyloseq object
```{r phyloseq_object, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

load('NomeDoProjeto_8.RData') # Load the previously saved objects here (item: 12. Importing the ready-made metadata table) to continue the analysis.

# Producing the phyloseq object
ps <- phyloseq(tax_table(taxtab),
               sample_data(samdf),
               otu_table(seqtab, taxa_are_rows = FALSE))
ps
saveRDS(ps, file = "ps.rds") # Saving the raw ps object. In this object, the sequence name will be the sequence itself.

# Renaming the sequences for better handling
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps.dna <- merge_phyloseq(ps, dna)
taxa_names(ps.dna) <- paste0("ASV", seq(ntaxa(ps.dna)))
ps.dna

saveRDS(ps.dna, file = "ps.dna.rds") # In this object, the sequence name will be ASV and a corresponding number. This object will be the usual one for the analysis. Throughout the analysis, for figures where it is necessary to format the names with the taxonomy, the 'format_to_best_hit' function should be used, which will possibly already be set in the chunk. This formatting cannot be done beforehand because some analysis functions do not tolerate this formatting. 
 
ps <- readRDS("ps.rds")
ps.dna <- readRDS("ps.dna.rds")

# Saving only the phyloseq objects
save(file = 'NomeDoProjeto_9.RData',list = c('ps', 'ps.dna'))

# Generating abundance table + taxonomy in CSV file

# Extracting abundance table
otu <- as.data.frame(otu_table(ps))
# Ensuring that ASVs are on the lines
if (!taxa_are_rows(ps)) {
  otu <- t(otu)
}
# Extract taxonomic table
tax <- as.data.frame(tax_table(ps))
# Combine abundance and taxonomy data
ASV_Table <- cbind(tax, otu)
# Rename the ASV column (Transform row names (ASV) into a column)
ASV_Table <- cbind(ASV = rownames(ASV_Table), ASV_Table)
rownames(ASV_Table) <- NULL
# Generates the CSV file
write.csv(ASV_Table, file = file.path(Preprocessing,"ASV_Table.csv"), row.names = FALSE, quote = FALSE)

```

## 15. Session info
```{r session_info, eval=TRUE, echo=TRUE}

sessionInfo()

```
**NomeDoProjeto_9.RData** Record: this object contains the output of phyloseq for further analysis.
