---
title: 'In this field, insert the project title'
author: 'In this field, insert the project title'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends 
---
================================================================================
# Study description:

~ Include here a brief description of the study ~


================================================================================

# PROCESSING - ALPHA AND BETA DIVERSITY ANALYSIS IN CROSS SECTIONAL STUDIES

# 1. Load libraries
```{r libraries_overview, eval=TRUE, echo=FALSE, warning=FALSE, include=FALSE}

library(phyloseq)
library(ggplot2)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(forcats)
library(writexl)   
library(grid)
library(here)

```

# 2. Data preparation

## 2.1. Plot Data Preparation
```{r data_prep_plot, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4,  fig.cap='Prevalence vs. Total Abundance (ASVs) with filtering cutoff'}

path         <- here::here() 
plots_dir <- file.path(path, 'prev_abund_plots')
dir.create(plots_dir)

# Load data
ps.dna <- readRDS("ps.dna.rds")
ps0.dna <- subset_taxa(ps.dna,
                       !is.na(Phylum) &
                       !Phylum %in% c("", "uncharacterized", "NA"))

prevalenceThreshold <- 0.1 * nsamples(ps0.dna) # Set prevalence threshold

# Prevalence x Abundance
prev_counts <- apply(otu_table(ps0.dna) > 0,
                     ifelse(taxa_are_rows(ps0.dna), 1, 2),
                     sum)

abund_counts <- taxa_sums(ps0.dna)

prevdf_asv <- data.frame(Prevalence = prev_counts,
                         Abundance  = abund_counts)

prev_plot <- ggplot(prevdf_asv, aes(x = Prevalence, y = Abundance)) +
  geom_point(alpha = 0.4) +
  geom_vline(xintercept = prevalenceThreshold,
             linetype = "dashed", color = "red3") +
  labs(
    subtitle = paste0("Filtering cutoff = ", prevalenceThreshold, " samples"),
    x        = "Number of samples in which each ASV appears",
    y        = "Total read count"
  ) +
  theme_bw(base_size = 12)

prev_plot

# Save
ggsave(filename = "prevalence_vs_abundance.png" , plot = prev_plot, path = plots_dir, width = 6, height = 4, units = "in", dpi = 300)
ggsave(filename = "prevalence_vs_abundance.pdf" , plot = prev_plot, path = plots_dir, width = 6, height = 4, units = "in", dpi = 300)
saveRDS(ps0.dna, "ps0.dna.rds")

# Clear env
rm(list = ls(all = TRUE))
  
```
\FloatBarrier

### 2.2. Table Data Preparation 
```{r data_prep_table, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

# Load data
ps0 <- readRDS("ps0.dna.rds")

# Filtering based on prevalence and abundance
prevalenceThreshold <- 0.1 * nsamples(ps0) # Set prevalence threshold based on section 2.1 findings
totalReads          <- sum(taxa_sums(ps0))
abundanceThreshold  <- 0.001 * totalReads          # 0.01%

# ASV
keepASV   <- taxa_sums(ps0) >= abundanceThreshold &
             apply(otu_table(ps0) > 0,
                   ifelse(taxa_are_rows(ps0), 1, 2),
                   sum) >= prevalenceThreshold

ps1  <- prune_taxa(keepASV, ps0)

# Phylum
ps0.phyla      <- tax_glom(ps1, "Phylum", NArm = FALSE)
keepPhylum     <- taxa_sums(ps0.phyla) >= abundanceThreshold &
                  apply(otu_table(ps0.phyla) > 0,
                        ifelse(taxa_are_rows(ps0.phyla), 1, 2),
                        sum) >= prevalenceThreshold
ps1.dna.phy    <- prune_taxa(keepPhylum, ps0.phyla)

# Family
ps0.family     <- tax_glom(ps1, "Family", NArm = FALSE)
keepFamily     <- taxa_sums(ps0.family) >= abundanceThreshold &
                  apply(otu_table(ps0.family) > 0,
                        ifelse(taxa_are_rows(ps0.family), 1, 2),
                        sum) >= prevalenceThreshold
ps1.dna.family <- prune_taxa(keepFamily, ps0.family)

# Genus
ps0.genus      <- tax_glom(ps1, "Genus",  NArm = FALSE)
keepGenus      <- taxa_sums(ps0.genus) >= abundanceThreshold &
                  apply(otu_table(ps0.genus) > 0,
                        ifelse(taxa_are_rows(ps0.genus), 1, 2),
                        sum) >= prevalenceThreshold
ps1.dna.genus  <- prune_taxa(keepGenus, ps0.genus)

# Table
counts <- tibble(
  Level      = c("ASV", "Genus", "Family", "Phylum"),
  Before     = c(ntaxa(ps0), ntaxa(ps0.phyla),
                 ntaxa(ps0.family), ntaxa(ps0.genus)),
  After      = c(ntaxa(ps1), ntaxa(ps1.dna.phy),
                 ntaxa(ps1.dna.family),ntaxa(ps1.dna.genus)),
  `Prev Thres`  = round(prevalenceThreshold, 1),
  `Abund Thres` = round(abundanceThreshold, 0)
)

# Format table for printing
counts %>%
  kable(
    caption = "Taxa before/after prevalence \\& abundance filtering",
    digits  = 0,
    booktabs= TRUE
  ) %>%
  kable_classic(full_width = FALSE, position = "center") %>%
  column_spec(2:3, width = "1.5cm") %>%
  column_spec(4:5, width = "1.2cm")


# Save
saveRDS(ps1,"ps1.rds")

# Clear environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

# 3. Microbiome overview

## 3.1. Microbiome sequencing depth
```{r dataSummary, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

path         <- here::here()
overview_dir <- file.path(path, 'overview')
dir.create(overview_dir)

# Load data
ps2    <- readRDS(file.path(path, 'ps1.rds'))

# Agglomerate by taxonomic level
ps2_phylum <- tax_glom(ps2, 'Phylum', NArm = TRUE)
ps2_family <- tax_glom(ps2, 'Family', NArm = TRUE)
ps2_genus  <- tax_glom(ps2, 'Genus',  NArm = TRUE)

# Metadata // Related info
# Adjust accordingly group names and factors (categorical variables)
meta <- data.frame(sample_data(ps2)) %>%
  mutate(
    Sample    = sample_names(ps2),
    Depth     = sample_sums(ps2),
    Group     = factor(Group,
                  levels = c("Group1",   
                             "Group2")),
    Sex       = factor(Sex),
    Treatment = factor(Treatment)
  )

# One-way ANOVA 
depth_aov <- aov(Depth ~ Group, data = meta)
anova_tbl <- broom::tidy(depth_aov)
colnames(anova_tbl)[1:6] <- c('Term','Df','Sum Sq','Mean Sq','F value','Pr(>F)')

anova_tbl %>%
  kable(
    caption    = 'One-way ANOVA on Sequencing Depth by Group',
    digits     = 3,
    booktabs   = TRUE
  ) %>%
  kable_styling(
    latex_options = c('hold_position','scale_down')
  ) %>%
  print()

```
\FloatBarrier

```{r depthSeq, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# Sequencing depth boxplot
p0 <- ggplot(meta, aes(x = Group, y = Depth, fill = Group)) + 
  geom_boxplot(width = 0.6) +
  scale_fill_manual(values = c(
    'Group 1' = 'steelblue', 
    'Group 2' = 'orange'     
  )) +
  labs(
    title = 'Sequencing depth by groups',
    x     = 'Group',
    y     = 'Sequencing depth (total number of reads)'
  ) +
  theme_bw(base_size = 8) +
  theme(legend.position = 'none')

# Save
ggsave(
  file.path(overview_dir, 'fig0_depth_boxplot.pdf'),
  p0, width = 180, height = 120, units = 'mm', dpi = 300
)

p0

```

```{r librarySize, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# Sample Library size 
p1 <- ggplot(meta, aes(x = Sample, y = Depth, fill = Group)) +
  geom_col() +
  scale_fill_manual(values = c(
    'Group 1'     = 'steelblue', # adjust accordingly
    'Group 2'     = 'orange'     
  )) +
  labs(
    x     = 'Sample ID',
    y     = 'Sequencing depth (total number of reads)',
    fill  = 'Group'
  ) +
  theme_bw(base_size = 8) +
  theme(
    axis.text.x   = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = 'top'
  ) 

# Save
  ggsave(
  file.path(overview_dir, 'fig1_library_size.pdf'),
  p1, width = 180, height = 120, units = 'mm', dpi = 300
  )

  p1
  
```
\FloatBarrier

## 3.2. Composition of microbiome
```{r funcRelat_abundance, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4}

# Function for relative abundance 
taxa_plot <- function(ps_obj, level_col, top_n, fig_num) {
  ps_ra <- transform_sample_counts(ps_obj, function(x) x / sum(x))
  df    <- psmelt(ps_ra)
  if (level_col == 'ASV') df <- rename(df, ASV = OTU)
  real_col <- if (level_col == 'ASV') 'ASV' else level_col
  top_taxa <- df %>%
    group_by(.data[[real_col]]) %>%
    summarize(mean_abund = mean(Abundance), .groups = 'drop') %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = top_n) %>%
    pull(real_col)
  df_plot <- df %>%
    mutate(
      Taxa = as.character(.data[[real_col]]),
      Taxa = ifelse(Taxa %in% top_taxa, Taxa, 'Other')
    ) %>%
    group_by(Sample, Group, Taxa) %>%
    summarize(Abundance = sum(Abundance), .groups = 'drop') %>%
    mutate(Taxa = fct_reorder(Taxa, Abundance, .fun = mean))

  p <- ggplot(df_plot, aes(x = Sample, y = Abundance, fill = Taxa)) +
    geom_bar(stat = 'identity') +
    facet_wrap(vars(Group), scales = 'free_x', ncol = 1) +
    labs(
      x     = 'Sample ID',
      y     = 'Relative Abundance',
      fill  = level_col
    ) +
    theme_bw(base_size = 8) +
    theme(
      axis.text.x       = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
      strip.background  = element_rect(fill = 'white'),
      strip.text        = element_text(face = 'bold'),
      legend.position   = 'right'
    )
  
  fname <- paste0('fig', fig_num, '_', tolower(level_col), '.pdf')
  
  ggsave(file.path(overview_dir, fname), p,
         width = 180, height = 120, units = 'mm', dpi = 300)
  print(p)
}

```

```{r f4, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at phylum level"}
taxa_plot(ps2_phylum, 'Phylum', 10, '2')

```

```{r f5, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at family level"}
taxa_plot(ps2_family, 'Family', 10, '3')

```

```{r f6, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at genus level"}
taxa_plot(ps2_genus, 'Genus', 10, '2')

```

```{r f7, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= "Relative abundance at ASV level"}
taxa_plot(ps2, 'ASV', 10, '2')

```
\FloatBarrier

## 3.3. General Dominant taxa
Here we transform the data in relative abundances for proper evaluation.
```{r microbiome_overview, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here::here() 
overview = paste(path,"/overview/",sep = "")

# Load data
ps0.dna <- readRDS("ps1.rds")

# Agglomerate
ps.dna.phy <- tax_glom(ps0.dna, "Phylum", NArm = TRUE)
ps.dna.family <- tax_glom(ps0.dna, "Family", NArm = TRUE)
ps.dna.genus <- tax_glom(ps0.dna, "Genus", NArm = TRUE)

# ASV level
ps.ra <- transform_sample_counts(ps0.dna, function(x) x/sum(x))

# Phylum level
taxa_names(ps.dna.phy) <- tax_table(ps.dna.phy)[,2]
ps.phy.ra <- transform_sample_counts(ps.dna.phy, function(x) x/sum(x))

# Family level
taxa_names(ps.dna.family) <- tax_table(ps.dna.family)[,5]
ps.family.ra <- transform_sample_counts(ps.dna.family, function(x) x/sum(x))

# Genus level
# taxa_names(ps.dna.genus) <- tax_table(ps.dna.genus)[,6]
ps.genus.ra <- transform_sample_counts(ps.dna.genus, function(x) x/sum(x))
genus.melt <- psmelt(ps.genus.ra)

```

The distribution of the reads are here summarized on phylum, family and genus level. 
```{r ov_microbiome_tables, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

df2 <- data.frame(tax_table(ps.dna.phy), taxprc = 100*taxa_sums(ps.phy.ra)/length(sample_names(ps.phy.ra)))
df3 <- data.frame(tax_table(ps.dna.family), taxprc = 100*taxa_sums(ps.family.ra)/length(sample_names(ps.family.ra)))
df4 <- data.frame(tax_table(ps.dna.genus),taxprc = 100*taxa_sums(ps.genus.ra)/length(sample_names(ps.genus.ra)))
df5 <- data.frame(tax_table(ps0.dna),taxprc = 100*taxa_sums(ps.ra)/length(sample_names(ps.ra)))

df.count <- data.frame(Included = c("All", "> 0.01%", "> 0.1%","> 1%"),
                          Phylum = c(nrow(df2),sum(df2$taxprc > 0.01),sum(df2$taxprc > 0.1),sum(df2$taxprc > 1)),
                          Family = c(nrow(df3),sum(df3$taxprc > 0.01),sum(df3$taxprc > 0.1),sum(df3$taxprc > 1)),
                          Genus = c(nrow(df4),sum(df4$taxprc > 0.01),sum(df4$taxprc > 0.1),sum(df4$taxprc > 1)),
                          ASV = c(nrow(df5),sum(df5$taxprc > 0.01),sum(df5$taxprc > 0.1),sum(df5$taxprc > 1)))

# Output file - Export to excel 

# abundance df for different taxonomic levels
FileName <- paste(overview,"/Abundance of phyla, family, genera and ASV in microbiome samples.xlsx", sep = "")
write_xlsx(df.count,FileName)

# abundance genus level df
FileName <- paste(overview,"/Average abundance according to genus.xlsx", sep = "")
write_xlsx(df4,FileName)

# abundance family level df
FileName <- paste(overview,"/Average abundance according to family.xlsx", sep = "")
write_xlsx(df3,FileName)

# abundance phylum level df
FileName <- paste(overview,"/Average abundance according to phylum.xlsx", sep = "")
write_xlsx(df2,FileName)

# Count of ASV and taxa in samples
kable(df.count, row.names = F,digits = 1, caption = 'Abundance of phyla, family, genera and ASV in microbiome samples')%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Top 6 dominating phyla (prc abundance)
kable(head(df2[order(df2$taxprc,decreasing = T),c("Kingdom","Phylum","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to phylum',col.names = c("Kingdom","Phylum","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Top 6 dominating family (prc abundance)
kable(head(df3[order(df3$taxprc,decreasing = T),c("Kingdom","Phylum","Family", "taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to family',col.names = c("Kingdom","Phylum", "Family", "Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Top 6 dominating (genera prc abundance)
kable(head(df4[order(df4$taxprc,decreasing = T),c("Kingdom","Phylum","Class","Order","Family","Genus","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to genus',col.names = c("Kingdom","Phylum","Class","Order","Family","Genus","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# clear environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

## 3.4. Distribution of taxa in the samples of Group 1
```{r group1_dist1, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here:here()
over_group = paste(path,"/overview_by_group/",sep = "")
dir.create(over_group)

# Load data
ps <- readRDS("ps1.rds")
ps.group1 <- subset_samples(ps, Group == 'Group 1') # Adjust group and variable name accordingly

# Agglomerating 
ps.group1.phy <- tax_glom(ps.group1, "Phylum", NArm = TRUE)
ps.group1.family <- tax_glom(ps.group1, "Family", NArm = TRUE)
ps.group1.genus <- tax_glom(ps.group1, "Genus", NArm = TRUE)

# ASV level
ps.group1.ra <- transform_sample_counts(ps.group1, function(x) x/sum(x))

# Phylum level
taxa_names(ps.group1.phy) <- tax_table(ps.group1.phy)[,2]
ps.group1.phy.ra <- transform_sample_counts(ps.group1.phy, function(x) x/sum(x))

# Family level
taxa_names(ps.group1.family) <- tax_table(ps.group1.family)[,5]
ps.group1.family.ra <- transform_sample_counts(ps.group1.family, function(x) x/sum(x))

# Genus level
# taxa_names(ps.dna.genus) <- tax_table(ps.dna.genus)[,6]
ps.group1.genus.ra <- transform_sample_counts(ps.group1.genus, function(x) x/sum(x))
ps.group1.genus.melt <- psmelt(ps.group1.genus.ra)

# Calculating relative abundance
df2 <- data.frame(tax_table(ps.group1.phy), taxprc = 100*taxa_sums(ps.group1.phy.ra)/length(sample_names(ps.group1.phy.ra)))
df3 <- data.frame(tax_table(ps.group1.family), taxprc = 100*taxa_sums(ps.group1.family.ra)/length(sample_names(ps.group1.family.ra)))
df4 <- data.frame(tax_table(ps.group1.genus),taxprc = 100*taxa_sums(ps.group1.genus.ra)/length(sample_names(ps.group1.genus.ra)))
df5 <- data.frame(tax_table(ps.group1),taxprc = 100*taxa_sums(ps.group1.ra)/length(sample_names(ps.group1.ra)))

df.count <- data.frame(Included = c("All", "> 0.01%", "> 0.1%","> 1%"),
                          Phylum = c(nrow(df2),sum(df2$taxprc > 0.01),sum(df2$taxprc > 0.1),sum(df2$taxprc > 1)),
                          Family = c(nrow(df3),sum(df3$taxprc > 0.01),sum(df3$taxprc > 0.1),sum(df3$taxprc > 1)),
                          Genus = c(nrow(df4),sum(df4$taxprc > 0.01),sum(df4$taxprc > 0.1),sum(df4$taxprc > 1)),
                          ASV = c(nrow(df5),sum(df5$taxprc > 0.01),sum(df5$taxprc > 0.1),sum(df5$taxprc > 1)))

# Count of ASV and taxa in group 1 samples
kable(df.count, row.names = F,digits = 1, caption = 'Abundance of phyla, family, genera and ASV in Group 1 samples')%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Abundance of phyla, family, genera and ASV in Group 1 samples.xlsx", sep = "")
write_xlsx(df.count, FileName)

# Top 6 dominating phyla (prc abundance)
kable(head(df2[order(df2$taxprc,decreasing = T),c("Kingdom","Phylum","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to phylum in Group 1 samples',col.names = c("Kingdom","Phylum","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Average abundance according to phylum in Group 1 samples.xlsx", sep = "")
write_xlsx(df2, FileName)

# Top 6 dominating family (prc abundance)
kable(head(df3[order(df3$taxprc,decreasing = T),c("Kingdom","Phylum","Family", "taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to family in Group 1 samples',col.names = c("Kingdom","Phylum", "Family", "Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Average abundance according to family in Group 1 samples.xlsx", sep = "")
write_xlsx(df3, FileName)

# Top 6 dominating (genera prc abundance)
kable(head(df4[order(df4$taxprc,decreasing = T),c("Kingdom","Phylum","Class","Order","Family","Genus","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to genus in Group 1 samples',col.names = c("Kingdom","Phylum","Class","Order","Family","Genus","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Average abundance according to genus in Group 1 samples.xlsx", sep = "")
write_xlsx(df4, FileName)

# clear environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

## 3.5. Distribution of taxa in the samples of Group 2
```{r group_2_dist1, eval=T, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here::here()
over_group = paste(path,"/overview_by_group/",sep = "")
dir.create(over_group)

# Load data
ps <- readRDS("ps1.rds")
ps.group2 <- subset_samples(ps, Group =='Group 2')

# Aglomerate
ps.group2.phy <- tax_glom(ps.group2, "Phylum", NArm = TRUE)
ps.group2.family <- tax_glom(ps.group2, "Family", NArm = TRUE)
ps.group2.genus <- tax_glom(ps.group2, "Genus", NArm = TRUE)

# ASV level
ps.group2.ra <- transform_sample_counts(ps.group2, function(x) x/sum(x))

# Phylum level
taxa_names(ps.group2.phy) <- tax_table(ps.group2.phy)[,2]
ps.group2.phy.ra <- transform_sample_counts(ps.group2.phy, function(x) x/sum(x))

# Family level
taxa_names(ps.group2.family) <- tax_table(ps.group2.family)[,5]
ps.group2.family.ra <- transform_sample_counts(ps.group2.family, function(x) x/sum(x))

# Genus level
# taxa_names(ps.dna.genus) <- tax_table(ps.dna.genus)[,6]
ps.group2.genus.ra <- transform_sample_counts(ps.group2.genus, function(x) x/sum(x))
ps.group2.genus.melt <- psmelt(ps.group2.genus.ra)

df2 <- data.frame(tax_table(ps.group2.phy), taxprc = 100*taxa_sums(ps.group2.phy.ra)/length(sample_names(ps.group2.phy.ra)))
df3 <- data.frame(tax_table(ps.group2.family), taxprc = 100*taxa_sums(ps.group2.family.ra)/length(sample_names(ps.group2.family.ra)))
df4 <- data.frame(tax_table(ps.group2.genus),taxprc = 100*taxa_sums(ps.group2.genus.ra)/length(sample_names(ps.group2.genus.ra)))
df5 <- data.frame(tax_table(ps.group2),taxprc = 100*taxa_sums(ps.group2.ra)/length(sample_names(ps.group2.ra)))

df.count <- data.frame(Included = c("All", "> 0.01%", "> 0.1%","> 1%"),
                          Phylum = c(nrow(df2),sum(df2$taxprc > 0.01),sum(df2$taxprc > 0.1),sum(df2$taxprc > 1)),
                          Family = c(nrow(df3),sum(df3$taxprc > 0.01),sum(df3$taxprc > 0.1),sum(df3$taxprc > 1)),
                          Genus = c(nrow(df4),sum(df4$taxprc > 0.01),sum(df4$taxprc > 0.1),sum(df4$taxprc > 1)),
                          ASV = c(nrow(df5),sum(df5$taxprc > 0.01),sum(df5$taxprc > 0.1),sum(df5$taxprc > 1)))

# Count of ASV and taxa in Group 2 samples
kable(df.count, row.names = F,digits = 1, caption = 'Abundance of phyla, family, genera and ASV in Group 2 samples')%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Abundance of phyla, family, genera and ASV in Group 2 samples.xlsx", sep = "")
write_xlsx(df.count, FileName)

# Top 6 dominating phyla (prc abundance)
kable(head(df2[order(df2$taxprc,decreasing = T),c("Kingdom","Phylum","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to phylum in Group 2 samples',col.names = c("Kingdom","Phylum","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Average abundance according to phylum in Group 2 samples.xlsx", sep = "")
write_xlsx(df2, FileName)

# Top 6 dominating family (prc abundance)
kable(head(df3[order(df3$taxprc,decreasing = T),c("Kingdom","Phylum","Family", "taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to family in Group 2 samples',col.names = c("Kingdom","Phylum", "Family", "Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Average abundance according to family in Group 2 samples.xlsx", sep = "")
write_xlsx(df3, FileName)

# Top 6 dominating (genera prc abundance)
kable(head(df4[order(df4$taxprc,decreasing = T),c("Kingdom","Phylum","Class","Order","Family","Genus","taxprc")]), row.names = F,digits = 1, caption = 'Average abundance according to genus in Group 2 samples',col.names = c("Kingdom","Phylum","Class","Order","Family","Genus","Abundance (%)"))%>% 
  kable_classic(full_width = F, position = "left")%>%  
  kable_styling(latex_options = c("scale_down", "hold_position"))

# Export to excel
FileName <- paste(over_group,"/Average abundance according to genus in Group 2 samples.xlsx", sep = "")
write_xlsx(df4,FileName)

# clear environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

# 4. Groups composition 
## 4.1. Top 10 Phyla
```{r phylum_abundance_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Top 10 phylum abundance by group'}

# Paths
path <- here::here()  
overview_group <- file.path(path, 'overview_by_group')
dir.create(overview_group, showWarnings = FALSE, recursive = TRUE)

# Load data
ps_ASV    <- readRDS(file.path(path, 'ps1.rds'))
ps_phy <- tax_glom(ps_ASV, 'Phylum', NArm = TRUE)

# Calculating relative abundance
ps_ra      <- transform_sample_counts(ps_phy, function(x) x / sum(x))
ps_ra_melt <- psmelt(ps_ra)

# Color palette
palette <- c(
  'Group 1'     = 'steelblue',
  'Group 2'     = 'orange'
)

# Top 10 Phyla per group
Phylum.summary <- ps_ra_melt %>%
  group_by(Group, Phylum) %>%
  summarize(mean_abund = mean(Abundance), .groups = 'drop') %>%
  arrange(desc(mean_abund))

top10 <- Phylum.summary %>%
  group_by(Group) %>%
  pull(Phylum) %>%
  unique() %>%
  head(n = 10)

df_plot <- ps_ra_melt %>%
  filter(Phylum %in% top10) %>%
  mutate(
    Abundance_percent = ifelse(Abundance < 0.001, 0.1, Abundance * 100),
    Phylum            = factor(Phylum, levels = rev(unique(Phylum)))
  )

n_taxa <- length(unique(df_plot$Phylum))

# Boxplot
F1 <- ggplot(df_plot, aes(x = Abundance_percent, y = Phylum)) +
  geom_boxplot(aes(color = Group), 
               position = position_dodge2(reverse = TRUE, width = 0.8),
               alpha = 0.5, outlier.shape = NA) +
  geom_boxplot(aes(fill = Group), 
               position = position_dodge2(reverse = TRUE, width = 0.8),
               alpha = 0.5, outlier.shape = 21) +
  scale_x_log10() +
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(
    x = "Abundance (%)",
    y = NULL
  ) +
  theme_bw(base_size = 8) +
  theme(
    axis.text.x      = element_text(size = 6),
    axis.text.y      = element_text(face = "italic", size = 6),
    axis.title       = element_text(size = 8),
    plot.title       = element_text(size = 9, hjust = 0.5),
    legend.position  = 'bottom',
    plot.margin      = unit(c(1,1,1,1), "lines")
  )

print(F1)

# Save
ggsave(
  file.path(overview_group, 'Top_Phylum_Abundance_by_Group.pdf'),
  F1, width = 180, height = 170, units = 'mm', dpi = 300
)
ggsave(
  file.path(overview_group, 'Top_Phylum_Abundance_by_Group.png'),
  F1, width = 180, height = 170, units = 'mm', dpi = 300
)

# Export to excel
write_xlsx(
  df_plot,
  file.path(overview_group, 'Top_Phylum_Abundance_by_Group.xlsx')
)

# Clear environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

## 4.2. Top 10 Family
```{r family_abundance_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Top 10 family abundance by group'}
  
path <- here::here()
overview_group <- file.path(path, 'overview_by_group')
dir.create(overview_group, showWarnings = FALSE, recursive = TRUE)

# Load data
ps_ASV    <- readRDS(file.path(path, 'ps1.rds'))
ps_family <- tax_glom(ps_ASV, 'Family', NArm = TRUE)

# Relative abundance
ps_ra      <- transform_sample_counts(ps_family, function(x) x / sum(x))
ps_ra_melt <- psmelt(ps_ra)

# Color palette
palette <- c(
  'Group 1'     = 'steelblue',
  'Group 2'     = 'orange'
)

# Top 10 Families per group
Family.summary <- ps_ra_melt %>%
  group_by(Group, Family) %>%
  summarize(mean_abund = mean(Abundance), .groups = 'drop') %>%
  arrange(desc(mean_abund))

top10 <- Family.summary %>%
  group_by(Group) %>%
  pull(Family) %>%
  unique() %>%
  head(n = 10) 

df_plot <- ps_ra_melt %>%
  filter(Family %in% top10) %>%
  mutate(
    Abundance_percent = ifelse(Abundance < 0.001, 0.1, Abundance * 100),
    Family            = factor(Family, levels = rev(unique(Family)))
  )

n_taxa <- length(unique(df_plot$Family))

# Boxplot
F1 <- ggplot(df_plot, aes(x = Abundance_percent, y = Family)) +
  geom_boxplot(aes(color = Group), 
               position = position_dodge2(reverse = TRUE, width = 0.8),
               alpha = 0.5, outlier.shape = NA) +
  geom_boxplot(aes(fill = Group), 
               position = position_dodge2(reverse = TRUE, width = 0.8),
               alpha = 0.5, outlier.shape = 21) +
  scale_x_log10() +
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(
    x = "Abundance (%)",
    y = NULL
  ) +
  theme_bw(base_size = 8) +
  theme(
    axis.text.x      = element_text(size = 6),
    axis.text.y      = element_text(face = "italic", size = 6),
    axis.title       = element_text(size = 8),
    plot.title       = element_text(size = 9, hjust = 0.5),
    legend.position  = 'bottom',
    plot.margin      = unit(c(1,1,1,1), "lines")
  )

print(F1)

# Save
ggsave(
  file.path(overview_group, 'Top_Family_Abundance_by_Group.pdf'),
  F1, width = 180, height = 170, units = 'mm', dpi = 300
)
ggsave(
  file.path(overview_group, 'Top_Family_Abundance_by_Group.png'),
  F1, width = 180, height = 170, units = 'mm', dpi = 300
)

# Export to excel
write_xlsx(
  df_plot,
  file.path(overview_group, 'Top_Family_Abundance_by_Group.xlsx')
)

# Clear environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

## 4.3. Top 10 Genus
```{r genus_abundance_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Top 10 genus abundance by group'}

# Paths
path <- here::here() 
overview_group <- file.path(path, 'overview_by_group')
dir.create(overview_group, showWarnings = FALSE, recursive = TRUE)

# Load data
ps_ASV    <- readRDS(file.path(path, 'ps1.rds'))
ps_phy <- tax_glom(ps_ASV, 'Genus', NArm = TRUE)

# Calculating relative abundance
ps_ra      <- transform_sample_counts(ps_phy, function(x) x / sum(x))
ps_ra_melt <- psmelt(ps_ra)

# Color palette
palette <- c(
  'Group 1'     = 'steelblue',
  'Group 2'     = 'orange'
)

# Top 10 per genus
Genus.summary <- ps_ra_melt %>%
  group_by(Group, Genus) %>%
  summarize(mean_abund = mean(Abundance), .groups = 'drop') %>%
  arrange(desc(mean_abund))

top10 <- Genus.summary %>%
  group_by(Group) %>%
  pull(Genus) %>%
  unique() %>%
  head(n = 10) 

df_plot <- ps_ra_melt %>%
  filter(Genus %in% top10) %>%
  mutate(
    Abundance_percent = ifelse(Abundance < 0.001, 0.1, Abundance * 100),
    Genus            = factor(Genus, levels = rev(unique(Genus)))
  )

n_taxa <- length(unique(df_plot$Genus))

# Boxplot
F1 <- ggplot(df_plot, aes(x = Abundance_percent, y = Genus)) +
  geom_boxplot(aes(color = Group), 
               position = position_dodge2(reverse = TRUE, width = 0.8),
               alpha = 0.5, outlier.shape = NA) +
  geom_boxplot(aes(fill = Group), 
               position = position_dodge2(reverse = TRUE, width = 0.8),
               alpha = 0.5, outlier.shape = 21) +
  scale_x_log10() +
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(
    x = "Abundance (%)",
    y = NULL
  ) +
  theme_bw(base_size = 8) +
  theme(
    axis.text.x      = element_text(size = 6),
    axis.text.y      = element_text(face = "italic", size = 6),
    axis.title       = element_text(size = 8),
    plot.title       = element_text(size = 9, hjust = 0.5),
    legend.position  = 'bottom',
    plot.margin      = unit(c(1,1,1,1), "lines")
  )

print(F1)

# Save
ggsave(
  file.path(overview_group, 'Top_Genus_Abundance_by_Group.pdf'),
  F1, width = 180, height = 170, units = 'mm', dpi = 300
)
ggsave(
  file.path(overview_group, 'Top_Genus_Abundance_by_Group.png'),
  F1, width = 180, height = 170, units = 'mm', dpi = 300
)

# Export to excel
write_xlsx(
  df_plot,
  file.path(overview_group, 'Top_Genus_Abundance_by_Group.xlsx')
)

# Clear environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

## 5. Session info
```{r session_info, eval=TRUE, echo=F}

sessionInfo()

```


