---
title: 'In this field, insert the project title'
author: 'In this field, insert the author(s) of the analysis'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends
---

# Study description:



# PROCESSING - ALPHA AND BETA DIVERSITY ANALYSIS IN CROSS SECTIONAL STUDIES

# 1. Load libraries
```{r loadLibraries, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

## Load libraries
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(forcats)
library(writexl)   
library(grid)
library(ggpubr) 
library(compositions)   
library(vegan)         
library(ape)
library(here)

```

# 2. Alpha diversity
## 2.1. Shannon index
```{r alpha_shannon_by_group, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Boxplot of alpha diversity – Shannon diversity index by Group',fig.align='center', fig.width=6, fig.height=4}

# Path
path      <- here::here()
alpha_div <- file.path(path, 'alpha_div')
dir.create(alpha_div, showWarnings = FALSE, recursive = TRUE)

# Phyloseq object
# ps1.rds is the phyloseq object in which ASVs were filtered based on prevalence and abundance. It is created in script #2.
ps <- readRDS(file.path(path, 'ps1.rds'))

#===============================================================================================================================

# Removing outliers, if necessary
#ps.no <- prune_samples(sample_names(ps) != "HIDRO(10)5", ps)

# If outliers are removed, remember to use the phyloseq object created below for further analysis
#saveRDS(ps.no, file.path(path, 'ps1.no_outlier.rds'))

#===============================================================================================================================

# color palette definition
palette <- c(
  'Group 1'     = 'steelblue',
  'Group 2'     = 'orange'
)

# Shannon index
alpha_df <- estimate_richness(ps, measures = "Shannon") %>%    # if there is an outlier: replace "ps" object > "ps.no".
  rownames_to_column(var = "Sample") %>%
  left_join(
    as(sample_data(ps), "data.frame") %>% rownames_to_column(var = "Sample"),
    by = "Sample"
  )

# Summary per group
alpha_sum <- alpha_df %>%
  group_by(Group) %>%
  summarise(
    `Samples (n)` = n(),
    Mean          = mean(Shannon),
    SD            = sd(Shannon),
    .groups       = 'drop'
  )

# Summary table
kable(
  alpha_sum,
  caption   = 'Summary of Shannon alpha diversity by Group',
  digits    = 3,
  col.names = c('Group', 'Samples (n)', 'Mean', 'SD')
) %>%
  kable_classic(full_width = FALSE, position = 'center') %>%
  add_header_above(c('', 'Shannon diversity index' = 3))

# Boxplot with Kruskal–Wallis p-value and legend
p_alpha <- ggplot(alpha_df, aes(x = Group, y = Shannon, fill = Group)) +
  geom_boxplot(alpha = 0.5, outlier.shape = 21,
               position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = palette, name = "Group") +
  stat_compare_means(
    aes(label = paste0("Kruskal–Wallis p = ", ..p.format..)),
    method  = "kruskal.test",
    label.x = 1.5,
    label.y = max(alpha_df$Shannon) * 1.05,
    size    = 3
  ) +
  labs(x = "Group", y = "Shannon diversity index") +
  theme_bw(base_size = 8) +
  theme(
    axis.title      = element_text(size = 10),
    axis.text       = element_text(size = 8),
    legend.position = "bottom",
    legend.title    = element_text(size = 8),
    legend.text     = element_text(size = 7)
  )

# Save
ggsave(
  file.path(alpha_div, 'alpha_shannon_by_group.pdf'),
  p_alpha, width = 180, height = 120, units = 'mm', dpi = 300
)
ggsave(
  file.path(alpha_div, 'alpha_shannon_by_group.png'),
  p_alpha, width = 180, height = 120, units = 'mm', dpi = 300
)

write_xlsx(
  alpha_df,
  file.path(alpha_div, 'alpha_shannon_data.xlsx')
)

print(p_alpha)

# Clear environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier
The differences for alpha diversity between groups were not statistically significant (Update accordingly the analisis).

# 3. Beta diversity 
## 3.1. Clustering - Techniques for dimensionality reduction and data visualization
### 3.1.1. PCoA (Principal Coordinate Analysis) Jaccard and Aitchison distances
```{r beta_div_ordination, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Path
path       <- here::here()
beta_div <- file.path(path, "beta_div")
dir.create(beta_div, showWarnings = FALSE, recursive = TRUE)
Clustering <- file.path(beta_div, "Clustering")
dir.create(Clustering, showWarnings = FALSE, recursive = TRUE)

# Load phyloseq object
ps <- readRDS(file.path(path, 'ps1.rds'))

# color palette definition
palette <- c(
  'Group 1'     = 'steelblue',
  'Group 2'     = 'orange'
)

# Metadata
meta <- sample_data(ps) %>%
  as("data.frame") %>%
  select(-Sample_ID) %>%  
  rownames_to_column("Sample_ID") %>%
  mutate(Group = factor(Group, levels = names(palette)))

# Build abundance df
seq_counts <- as.data.frame(t(otu_table(ps)))

```
\FloatBarrier

```{r beta_div_ordination_pcoaJac, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'PCoA (Jaccard) by group'}

# PCoA Jaccard
jac_dmat <- vegdist(t(seq_counts), method = "jaccard")
pcoa_res <- ape::pcoa(jac_dmat)
jvar     <- pcoa_res$values$Relative_eig
df_pcoa  <- data.frame(Axis.1 = pcoa_res$vectors[,1],
                       Axis.2 = pcoa_res$vectors[,2],
                       meta)

# PCoA plot Jaccard
p_pcoaJac <- ggplot(df_pcoa, aes(Axis.1, Axis.2, color = Group)) +
  stat_ellipse(size = 1, level = 0.95) +
  geom_point(size = 3) +
  scale_color_manual(values = palette, name = "Group") +
  coord_fixed(ratio = jvar[2]/jvar[1]) +
  labs(title = "PCoA - Jaccard distance",
    x     = paste0("Axis 1 (", round(jvar[1]*100,1), "%)"),
    y     = paste0("Axis 2 (", round(jvar[2]*100,1), "%)")
  ) +
  theme_bw(base_size = 8) +
  theme(legend.position = "right")

p_pcoaJac

# Save
ggsave(file.path(Clustering, "pcoa_jaccard_by_group.pdf"),
       p_pcoaJac, width = 180, height = 120, units = "mm", dpi = 300)

write_xlsx(df_pcoa, file.path(Clustering, "pcoa_jaccard_scores.xlsx"))

```

```{r beta_div_ordination_pcoaAitch, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'PCoA (Jaccard) by group'}

# PCoA Aitchison

set.seed(123) # Always run 

# Build abundance df
seq_counts <- as.data.frame(t(otu_table(ps)))

# add "pseudocount"
invisible(capture.output({
  seq_counts_noZero <- zCompositions::cmultRepl(seq_counts)})) # method Count Zero Multiplicative-CZM

# PCoA Aitchison's Euclidean
aitch_dmat <- vegan::vegdist(t(seq_counts_noZero), method = "aitchison")
pcoaAitch_res <- ape::pcoa(aitch_dmat)
jvar     <- pcoaAitch_res$values$Relative_eig
df_pcoaAitch  <- data.frame(Axis.1 = pcoaAitch_res$vectors[,1],
                       Axis.2 = pcoaAitch_res$vectors[,2],
                       meta)

# PCoA plot Aitchison
p_pcoa_Aitch <- ggplot(df_pcoaAitch, aes(Axis.1, Axis.2, color = Group)) +
  stat_ellipse(size = 1, level = 0.95) +
  geom_point(size = 3) +
  scale_color_manual(values = palette, name = "Group") +
  coord_fixed(ratio = jvar[2]/jvar[1]) +
  labs(title = "PCoA - Aitchison's Euclidean distance",
    x     = paste0("Axis 1 (", round(jvar[1]*100,1), "%)"),
    y     = paste0("Axis 2 (", round(jvar[2]*100,1), "%)")
  ) +
  theme_bw(base_size = 8) +
  theme(legend.position = "right")

p_pcoa_Aitch

# Save
ggsave(file.path(Clustering, "pcoa_aitchison_by_group.pdf"),
       p_pcoa_Aitch, width = 180, height = 120, units = "mm", dpi = 300)

write_xlsx(df_pcoaAitch, file.path(Clustering, "pcoa_aitchison_scores.xlsx"))

```
\FloatBarrier

### 3.1.2. NMDS (Non-Metric Multidimensional Scaling) Jaccard and Aitchison distances
```{r beta_div_ordination_nmds, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'NMDS - Jaccard distance'}

# NMDS (Jaccard and Aitchison)

set.seed(123) # Always run before NMDS

# Jaccard 
jac_nmds <- metaMDS(jac_dmat, k = 2, autotransform = FALSE, trymax = 100, trace = 0)
 
df_jac_nd <- data.frame(MDS1 = jac_nmds$points[,1],
                        MDS2 = jac_nmds$points[,2],
                        meta)

# Aitchison distances 
euc_nmds <- metaMDS(aitch_dmat, k = 3, autotransform = FALSE, trymax = 100, trace = 0)
df_euc_nd <- data.frame(MDS1 = euc_nmds$points[,1],
                        MDS2 = euc_nmds$points[,2],
                        meta)

# k value
k_value_jac <- jac_nmds$ndim 
k_value_euc <- euc_nmds$ndim 

```
\FloatBarrier

```{r beta_div_ordination_nmds_jac, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'NMDS - Jaccard distance'}

# plot NMDS Jaccard
p_nmds_jac <- ggplot(df_jac_nd, aes(MDS1, MDS2, color = Group)) +
  stat_ellipse(size = 1, level = 0.95) +
  geom_point(size = 3) +
  labs(title = paste0("NMDS - Jaccard (k = ", k_value_jac, "), Stress = ", round(jac_nmds$stress,3))) +
  scale_color_manual(values = palette, name = "Group") +
  theme_bw(base_size = 8) +
  theme(legend.position = "right")

p_nmds_jac

# Saving NMDS_jac

ggsave(file.path(Clustering, "nmds_jaccard_by_group.pdf"),
       p_nmds_jac, width = 180, height = 120, units = "mm", dpi = 300)
write_xlsx(df_jac_nd, file.path(Clustering, "nmds_jaccard_scores.xlsx"))

```
\FloatBarrier

```{r beta_div_ordination_nmds_ait, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'NMDS - Aitchison distance'}

# Plot NMDS Aitchison's Euclidean
p_nmds_euc <- ggplot(df_euc_nd, aes(MDS1, MDS2, color = Group)) +
  stat_ellipse(size = 1, level = 0.95) +
  geom_point(size = 3) +
  labs(title = paste0("NMDS Aitchison's Euclidean (k = ", k_value_euc, "), Stress = ", round(euc_nmds$stress, 3))) +
  scale_color_manual(values = palette, name = "Group") +
  theme_bw(base_size = 8) +
  theme(legend.position = "right")

p_nmds_euc

# Saving NMDS_Euclidean
ggsave(file.path(Clustering, "nmds_aitchison_by_group.pdf"),
       p_nmds_euc, width = 180, height = 120, units = "mm", dpi = 300)

write_xlsx(df_euc_nd, file.path(Clustering, "nmds_aitchison_scores.xlsx"))

```
\FloatBarrier
To corroborate the previous findings, we will perform an hierarchical clustering.

### 3.1.3. Hierarchical clustering - Jaccard distance
```{r hclust_jac, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap= 'Jaccard hierarchical agglomerative clustering'}

cluster_ex<-hclust(vegdist(t(seq_counts),method='jaccard'),method="complete")
plot(cluster_ex, xlab='',sub='')

# Save
png(paste0(Clustering, "/cluster_dendrogram_jaccard.png"), width = 800, height = 600) 
plot(cluster_ex, xlab='',sub='')
invisible(dev.off())

```
\FloatBarrier

## 3.2. Beta diversity and statistical tests
### 3.2.1. Beta dispersion
```{r beta_dispersion, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Path
stats_test.dir <- file.path(beta_div, "stats_test")
dir.create(stats_test.dir, showWarnings = FALSE, recursive = TRUE)

# β-dispersion: homogeneity of group variances

set.seed(123) # Always run this line before betadisper()

bd_ait   <- betadisper(aitch_dmat, meta$Group)
bd_jac   <- betadisper(jac_dmat, meta$Group)
disp_ait <- anova(bd_ait)
disp_jac <- anova(bd_jac)

disp_table <- tibble::tibble(
  Distance = c("Aitchison", "Jaccard"),
  F.value  = c(disp_ait$F[1],  disp_jac$F[1]),
  P.value  = c(disp_ait$Pr[1], disp_jac$Pr[1])
)

# Print summary
disp_table %>%
  knitr::kable(
    digits  = 3,
    caption = "Test for homogeneity of dispersion (betadisper)"
  ) %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"))

# Save
disp_table %>%
  knitr::kable(
    digits  = 3,
    caption = "Test for homogeneity of dispersion (betadisper)",
    format = "html"
  ) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::save_kable(file = file.path(stats_test.dir, "betadisper_results.html"))

```
\FloatBarrier

### 3.2.2. PERMANOVA for Aitchison distance and Group variable
```{r permanova_ait, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Output path
stats_test.dir <- file.path(beta_div, "stats_test")

# PERMANOVA 
df <- microbiome::meta(ps)

perm <- with(df, how(nperm = 999))
perm_ait_tp <- adonis2(aitch_dmat ~ Group, data = df, permutations = perm)

# Print summary
broom::tidy(perm_ait_tp) %>%
  knitr::kable(
    digits  = 3,
    caption = "PERMANOVA (Aitchison) – Group effect",
  ) %>%
  kableExtra::kable_styling(latex_options = c("hold_position","scale_down"))

# Save
broom::tidy(perm_ait_tp) %>%
  knitr::kable(
    digits  = 3,
    caption = "PERMANOVA (Aitchison) – Group effect",
    format = 'html'
  ) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::save_kable(file = file.path(stats_test.dir, "permanova_results.html"))

```
\FloatBarrier

### 3.2.3. PERMANOVA for Jaccard distance and Group variable
```{r permanova_jac, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Path
stats_test.dir <- file.path(beta_div, "stats_test")

# PERMANOVA for Jaccard distance test
perm <- with(df, how(nperm = 999))
perm_jac_tp <- adonis2(jac_dmat ~ Group, data = df, permutations = perm)

# Print summary
broom::tidy(perm_jac_tp) %>%
  knitr::kable(
    digits  = 3,
    caption = "PERMANOVA (Jaccard) – Group effect"
  ) %>%
  kableExtra::kable_styling(latex_options = c("hold_position","scale_down"))

# Save
broom::tidy(perm_jac_tp) %>%
  knitr::kable(
    digits  = 3,
    caption = "PERMANOVA (Jaccard) – Group effect",
    format = "html"
  ) %>%
  kableExtra::kable_styling(full_width = FALSE) %>%
  kableExtra::save_kable(file = file.path(stats_test.dir, "permanovaJaccard_results.html"))

# clear environment
#rm(list = ls(all = TRUE))

```
\FloatBarrier

# 4. Session info
```{r session_info, eval=TRUE, echo=F}

sessionInfo()

```


