---
title: 'In this field, insert the project title'
author: 'In this field, insert the author(s) of the analysis'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  word_document:
    toc: yes
    toc_depth: 5
  html_document:
    df_print: paged
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: yes
    code_folding: show
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}           # \FloatBarrier
  - \usepackage[font=small]{caption} # Legends 
---
================================================================================ 
# Study description:

~ Include here a brief description of the study ~

================================================================================

# DIFFERENTIAL ABUNDANCE ANALYSIS IN CROSS SECTIONAL STUDIES

## 1. Load libraries
```{r libraries_diffAbund, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

library(ANCOMBC)
library(tidyverse)
library(DT)
options(DT.options = list(
  initComplete = JS("function(settings, json) {",
  "$(this.api().table().header()).css({'background-color': 
  '#000', 'color': '#fff'});","}")))
library(writexl)
library(dplyr)
library(tidyr)
library(ggplot2)

```

## 2. Analysis of Compositions of Microbiomes with Bias Correction 2 (ANCOMBC2)
We performed the differential abundance analysis at genus level with ANCOMBC2
```{r run_ANCOMBC2, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Paths
path <- here::here()
Dif_abundance = paste(path,"/Differential_abundance_genus/",sep = "")
dir.create(Dif_abundance)

# Load phyloseq data 
ps <- readRDS("ps1.rds") # ps1.rds is the phyloseq object in which ASVs were filtered based on prevalence and abundance. It is created in script 2 (overview).

set.seed(123) # Always run this line before differential abundance analysis

# Run ANCOM-BC
output_genus = ancombc2(data = ps, assay_name = "counts", tax_level = "Genus", 
                  fix_formula = "Group", rand_formula = NULL, 
                  p_adj_method = "BH", pseudo_sens = TRUE, 
                  prv_cut = 0.1, lib_cut = 1000, s0_perc = 0.05, 
                  group = "Group", struc_zero = TRUE, neg_lb = TRUE, 
                  alpha = 0.05, n_cl = 20, verbose = TRUE,
                  global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = FALSE, 
                  iter_control = list(tol = 1e-2, max_iter = 20, 
                                      verbose = TRUE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100), 
                  trend_control = list(contrast = list(matrix(c(1, 0, -1, 1),
                                                              nrow = 2, 
                                                              byrow = TRUE),
                                                       matrix(c(-1, 0, 1, -1),
                                                              nrow = 2, 
                                                              byrow = TRUE),
                                                       matrix(c(1, 0, 1, -1),
                                                              nrow = 2, 
                                                              byrow = TRUE)),
                                       node = list(2, 2, 1),
                                       solver = "ECOS",
                                       B = 100))
# Save
save(file = "ANCOMBC.RData", list = "output_genus")

results <- output_genus$res

# Save results
write.csv(results, 
          file = "results_ancombc.csv", 
          row.names = FALSE) 

# clear environment
rm(list = ls(all = TRUE))

```

### 2.1. Genus level
```{r struc_zeros, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Path
path <- here::here()
Dif_abundance = paste(path,"/Differential_abundance_genus/",sep = "")

load('ANCOMBC.RData')

# Structural zeros 
tab_zero = output_genus$zero_ind
tab_zero %>%
    datatable(caption = "The detection of structural zeros on Genus level")
FileName <- paste(Dif_abundance,"/Structural_zeros_genus.xlsx", sep = "")
write_xlsx(tab_zero, FileName)

```

### 2.1.1. Overall - Primary results 
```{r ANCOMBC2_resPrim_heatmap, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Global differential abundance (ANCOM-BC2) – taxa significant in any group contrast', fig.align='center', fig.width=8, fig.height=11}

# Primary analysis 

# Path
path    <- here::here()
outdir  <- file.path(path, 'Differential_abundance_genus')
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

# Primary LFCs and global test results
res_prim   <- output_genus$res

df_group = res_prim %>%
    dplyr::select(taxon, contains("Group")) 

# Groups definition - IMPORTANT: Adjust variable names accordingly to the project
group_ref <- 'Group1' # [insert the identification of reference/baseline group to compare]
group_exp1 <- paste0('Group',' ', 2) # [insert here the identification of experimental group]

df = df_group %>%
    dplyr::filter(paste0("diff_Group", group_ref) == 1 | 
                  paste0("diff_Group", group_exp1) == 1 ) %>%
    dplyr::mutate(lfc1 = ifelse(paste0("diff_Group", group_ref) == 1, 
                                round(paste0("lfcGroup", group_ref), 2), 0),    
                  lfc2 = ifelse(paste0("diff_Group", group_exp1) == 1, 
                                round(paste0("lfcGroup", group_exp1), 2), 0)) %>%
    tidyr::pivot_longer(cols = lfc1:lfc2, 
                        names_to = "group", values_to = "value") %>%
    dplyr::arrange(taxon)

#=========================================================================================================================

# Plot
# Compute gradient limits
lo  <- floor(min(df$value, na.rm = TRUE))
up  <- ceiling(max(df$value, na.rm = TRUE))
mid <- (lo + up) / 2

# Horizontal bar plot
bar_global = df %>%
  ggplot(aes(y = reorder(taxon, abs(value)), x = value, 
             fill = ifelse(value >= 0, "Positive LFC", "Negative LFC"))) +
  geom_bar(stat = "identity", width = 0.5, color = "black") +
  labs(y = "", x = "Abundance", title = "") +
  scale_fill_manual(values = c("Positive LFC" = "darkslategray3", "Negative LFC" = "coral"),
                    name = "") +
  facet_wrap(~group, nrow = 3) +  
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0),
    axis.text.y = element_text(face = "italic", size = 7),
    axis.text.x = element_text(size = 10),
    strip.text.x = element_text(size = 8),
    plot.margin = unit(c(1, 1, 1, 1), "lines")
  )

bar_global # display the plot

# Save
ggsave(file.path(outdir, 'ANCOMBC2_primaryResults_barchart.pdf'),
       bar_global, width = 180, height = 200, units = "mm", dpi = 300)
ggsave(file.path(outdir, 'ANCOMBC2_primaryResults_barchart.png'),
       bar_global, width = 180, height = 200, units = "mm", dpi = 300)

# clear environment
rm(list = ls(all = TRUE))

```

### 2.1.2. Global results
```{r ANCOMBC2_global_heatmap, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Global differential abundance (ANCOM-BC2) – taxa significant in any group contrast', fig.align='center', fig.width=8, fig.height=11}

# Global results - If there are at least 3 experimental groups in the study, run the code below

load('ANCOMBC.RData')

# Path
path    <- here::here()
outdir  <- file.path(path, 'Differential_abundance_genus')
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

# Pull in primary LFCs and global test results
res_prim   <- output_genus$res
res_global <- output_genus$res_global

# Groups definition
group_ref  <- "Group 1" # [insert the identification of reference/baseline group to compare]
group_exp1 <- "Group 2" # [insert the identification of experimental group]
group_exp2 <- "Group 3"  # opcional 

# Join per‐group LFCs to the global diff_abn flag
df <- res_prim %>%
  select(taxon,
    lfc1 = paste0("lfc_Group", group_exp1),
    lfc2 = paste0("lfc_Group", group_exp2))
  ) %>%
  left_join(
    res_global %>% transmute( taxon,
                              diff_abn  = diff_abn,
                              passed_ss = passed_ss),
    by = "taxon"
  ) %>%
  # Keep only those taxa flagged in the global test:
  filter(diff_abn == TRUE) %>%
  # Build long form for values and colors
  dplyr::mutate(
    lfc1 = round(lfc1, 2),
    lfc2 = round(lfc2, 2),
    col1 = ifelse(passed_ss == TRUE, "aquamarine3", "black"),
    col2 = ifelse(passed_ss == TRUE, "aquamarine3", "black")
  ) %>%
  pivot_longer(cols = starts_with("lfc"),
               names_to  = "grp", values_to = "value") %>%
    pivot_longer(cols = starts_with("col"), 
                 names_to  = "grp2", values_to = "color") %>%
  filter(sub("lfc", "col", grp) == grp2) %>%
  select(taxon, grp, value, color) %>%
  # Group labels
  mutate(group = recode(grp,
      lfc1 = paste0(group_exp1, " – ", group_ref),
      lfc2 = paste0(group_exp2, " – ", group_ref)
    ),
    group = factor(group, levels = c(
      paste0(group_exp1, " – ", group_ref),
      paste0(group_exp2, " – ", group_ref)
    ))
  ) %>% 
  # Order taxa within each facet
  group_by(group) %>%
  arrange(value) %>%
  mutate(taxon = factor(taxon, levels = unique(taxon))) %>%
  ungroup()

#=========================================================================================================================

# Plot 
# Compute gradient limits
lo  <- floor(min(df$value, na.rm = TRUE))
up  <- ceiling(max(df$value, na.rm = TRUE))
mid <- (lo + up) / 2

# Horizontal bar plot
bar_global = df %>%
  ggplot(aes(y = reorder(taxon, abs(value)), x = value, 
             fill = ifelse(value >= 0, "Positive LFC", "Negative LFC"))) +
  geom_bar(stat = "identity", width = 0.5, color = "black") +
  labs(y = "", x = "Abundance", title = "") +
  scale_fill_manual(values = c("Positive LFC" = "darkslategray3", "Negative LFC" = "coral"),
                    name = "") +
  facet_wrap(~group, nrow = 3) +  
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0),
    axis.text.y = element_text(face = "italic", size = 7),
    axis.text.x = element_text(size = 10),
    strip.text.x = element_text(size = 8),
    plot.margin = unit(c(1, 1, 1, 1), "lines")
  )

bar_global # display the plot

# saving bar_global
ggsave(file.path(outdir, 'ANCOMBC2_global_barchart.pdf'),
       bar_global, width = 180, height = 200, units = "mm", dpi = 300)
ggsave(file.path(outdir, 'ANCOMBC2_global_barchart.png'),
       bar_global, width = 180, height = 200, units = "mm", dpi = 300)

# clean environment
rm(list = ls(all = TRUE))

```
\FloatBarrier

### 2.1.3. Pairwise results
```{r ANCOMBC2_pairwise_heatmap, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Pairwise differential abundance (ANCOM-BC2) – all group–group contrasts',fig.align='center', fig.width=8, fig.height=7}

# Pairwise comparison - If there are at least 3 experimental groups in the study, run the code below

load('ANCOMBC.RData')

# Get pairwise results
res_pair <- output_genus$res_pair

# Build df_fig_pair1: extract LFCs where diff==1
df_fig_pair1 <- res_pair %>% dplyr::filter(
  paste0("diff_Group", group_ref)  == 1|
  paste0("diff_Group", group_exp1) == 1 |
  paste0("diff_Group", group_exp2) == 1
  ) %>%
  dplyr::mutate(
    lfc1 = ifelse(paste0("diff_Group", group_ref) == 1,
      round(paste0("lfc_Group", group_ref), 2), 0),
    lfc2 = ifelse(paste0("diff_Group", group_exp1) == 1,
      round(paste0("lfc_Group", group_exp1), 2), 0),
    lfc3 = ifelse(paste0("diff_Group", group_exp2) == 1,
      round(paste0("lfc_Group", group_exp2), 2), 0)
  ) %>%
  tidyr::pivot_longer(
    cols      = lfc1:lfc3,
    names_to  = "group",
    values_to = "value"
  ) %>%
  dplyr::arrange(taxon)

# Build df_fig_pair2: extract colors where passed_ss & diff==1
df_fig_pair2 <- res_pair %>%
  dplyr::filter(
    !!rlang::sym(paste0("diff_Group", group_ref))  == 1 |
    !!rlang::sym(paste0("diff_Group", group_exp1)) == 1 |
    !!rlang::sym(paste0("diff_Group", group_exp2)) == 1
  ) %>%
  dplyr::mutate(
    lfc1 = ifelse(
      paste0("passed_ss_Group", group_ref) == 1 &
      paste0("diff_Group", group_ref) == 1,
      "aquamarine3", "black"),
    lfc2 = ifelse(
      paste0("passed_ss_Group", group_exp1) == 1 &
      paste0("diff_Group", group_exp1) == 1,
      "aquamarine3", "black"),
    lfc3 = ifelse(
      !!rlang::sym(paste0("passed_ss_Group", group_exp2)) == 1 &
      !!rlang::sym(paste0("diff_Group", group_exp2)) == 1,
      "aquamarine3", "black")
  ) %>%
  tidyr::pivot_longer(
    cols      = lfc1:lfc3,
    names_to  = "group",
    values_to = "color"
  ) %>%
  dplyr::arrange(taxon)

# Join values and colors
df_fig_pair <- df_fig_pair1 %>%
  dplyr::left_join(df_fig_pair2, by = c("taxon", "group"))

# Recode and factor the group labels
df_fig_pair$group <- dplyr::recode(df_fig_pair$group,
  lfc1 = paste0(group_ref,  " - baseline"),
  lfc2 = paste0(group_exp1, " - ", group_ref),
  lfc3 = paste0(group_exp2, " - ", group_ref)
)
df_fig_pair$group <- factor(df_fig_pair$group,
  levels = c(
    paste0(group_ref,  " - baseline"),
    paste0(group_exp1, " - ", group_ref),
    paste0(group_exp2, " - ", group_ref)
  )
)

#=========================================================================================================================

# Plot
# Compute gradient limits
lo  <- floor(min(df_fig_pair$value, na.rm = TRUE))
up  <- ceiling(max(df_fig_pair$value, na.rm = TRUE))
mid <- (lo + up) / 2

# Horizontal pairwise barplot
bar_pair2 = df_fig_pair %>%
  ggplot(aes(y = reorder(taxon, abs(value)), x = value, 
             fill = ifelse(value >= 0, "Positive LFC", "Negative LFC"))) +
  geom_bar(stat = "identity", width = 0.5, color = "black") +
  labs(y = "", x = "Abundance", title = "") +
  scale_fill_manual(values = c("Positive LFC" = "darkslategray3", "Negative LFC" = "coral"),
                    name = "") +
  facet_wrap(~group, nrow = 3) +  # <--- ADD the plus sign here
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0),
    axis.text.y = element_text(face = "italic", size = 7),
    axis.text.x = element_text(size = 10),
    strip.text.x = element_text(size = 8),
    plot.margin = unit(c(1, 1, 1, 1), "lines")
  )

bar_pair2 # Display the plot

# Save
ggsave(file.path(outdir, 'ANCOMBC2_pairwise_barchart2.pdf'),
       bar_pair2, width = 180, height = 170, units = "mm", dpi = 300)
ggsave(file.path(outdir, 'ANCOMBC2_pairwise_barchart2.png'),
       bar_pair2, width = 180, height = 170, units = "mm", dpi = 300)

# clear environment
rm(list = ls(all = TRUE))

```

## 3. Session info
```{r session_info, eval=TRUE, echo=FALSE}

sessionInfo()

```